%=======================================================================
%  Finite-Support Periodic Highways of Langton's Ant:
%  Necessary Conditions, Transverse Exclusions, and Exact Search
%
%  Author: Atharva Jillhewar
%
%  SELF-CONTAINED: paste this single file into Overleaf and press
%  "Recompile" (pdfLaTeX).  The bibliography is embedded, so there is NO
%  separate .bib file and NO BibTeX step required.  References and
%  appendices render in one compile (Overleaf runs the 2 pdfLaTeX passes
%  needed to resolve \cite and \ref automatically).
%
%  Author metadata is filled below; update it only if the author requests a change.
%=======================================================================
\documentclass[11pt,reqno]{amsart}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}      % Type 1 vector fonts: keeps the PDF text layer searchable/copyable
\usepackage{microtype}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{listings}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true,linkcolor=blue!55!black,citecolor=teal!60!black,
            urlcolor=purple!55!black,
            pdftitle={Finite-Support Periodic Highways of Langton's Ant: Necessary Conditions, Transverse Exclusions, and Exact Search},
            pdfauthor={Atharva Jillhewar}]{hyperref}

%------------------------------ theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{problem}[theorem]{Problem}

%------------------------------ macros
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\che}{\chi}
\newcommand{\drift}{d}
\newcommand{\growth}{g}
\newcommand{\turnR}{\mathrm{R}}
\newcommand{\turnL}{\mathrm{L}}
\newcommand{\vE}{\mathsf{E}}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\supp}{supp}

%------------------------------ code listing style (ASCII-faithful Lean 4)
\lstdefinestyle{lean}{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\bfseries,
  commentstyle=\itshape\color{gray},
  morekeywords={def,theorem,lemma,by,match,with,structure,namespace,end,
    import,fun,let,in,Bool,List,true,false,rw,simp,induction,cases,exact,
    refine,intro,rfl},
  showstringspaces=false,
  columns=fullflexible,
  breaklines=true,
  frame=single,
  framesep=5pt,
  rulecolor=\color{gray!40},
  backgroundcolor=\color{gray!5},
}

\title[Finite-support periodic highways]{Finite-Support Periodic Highways of Langton's Ant:\\ Necessary Conditions, Transverse Exclusions, and Exact Search}

\author{Atharva Jillhewar}
\address{Independent researcher}
\email{Atharvajil124@gmail.com}

\date{21 July 2026}

\keywords{Langton's ant, turmite, cellular automata, highway conjecture,
symbolic dynamics, reversible dynamics, medial graph, Tait graph, interlacement}
\subjclass[2020]{Primary 37B15; Secondary 37B10, 05C10, 68Q80, 68R15}

\begin{document}
\raggedbottom

\begin{abstract}
We study the classical two-colour Langton's ant on $\Z^2$, started from a
colouring with finite black support, under the additional hypothesis that its turn
trace is eventually periodic with nonzero drift. We give a necessary-and-sufficient,
decidable criterion for a finite word to have such a finite-support periodic
realisation, together with an explicit seed. We prove that every resulting highway
has strictly positive net black growth divisible by four, and derive a signed
mod-four residue identity for its growing wake and the bound
$\growth\ge2\max(|a|,|b|)$ for drift $(a,b)$. We also give an exact medial
(Tait)-graph formulation and a two-endpoint collision-chain parity identity. For
diagonal drift we prove a transverse rigidity theorem: the two extremal level lines
of the trace are horizontal, carry only the turn $\turnR$, and every cell of them
reached by the periodic tail is entered exactly once and never revisited within that
tail, so the highway is bounded on
both sides by guard rails of permanent wake; the transverse width is therefore even. From
this we deduce that no such highway has transverse width two.  We then give a
computer-assisted all-period exclusion of width four: exact crossing sequences at
successive untouched five-cell columns belong to a twelve-edge directed graph, and
every edge strictly lowers an explicit rank.  Two independently written local
enumerators reproduce the complete edge table, while Lean checks its rank
consequence.  Thus every diagonal periodic highway has transverse width at least
six. A
residue-theorem-free audited enumeration applies the realisability criterion
to every positive-growth, nonzero-drift leaf and excludes all periods at most $48$.
Selected finite algebraic kernels are checked in Lean~$4$; the geometric extraction
of those kernels from an ant trace is proved conventionally here but is not yet
formalised end to end. These results constrain periodic highways but prove neither
that every finite-support orbit eventually becomes periodic nor that the standard
period-$104$ highway is the unique positive-growth periodic trace.
\end{abstract}

\maketitle

%=======================================================================
\section{Introduction}
%=======================================================================

\emph{Langton's ant} is the turmite on the square lattice $\Z^2$ defined as
follows. Each cell is coloured white or black; a pointer, the \emph{ant}, occupies
a cell and faces one of the four cardinal directions. At each step the ant reads
the colour of its current cell: on white it turns $90^\circ$ right, on black it
turns $90^\circ$ left; it then flips the colour of the current cell and moves one
unit forward in its new heading. In the convention used here, direct simulation from
the all-white plane enters, after $9977$ steps, a diagonal translating
structure---the \emph{highway}---whose turn trace has least period $104$ and whose
ant position translates by a vector $(\pm2,\pm2)$ each period. The rule was
introduced by Langton \cite{Langton1986}; the exact phase used here is reproduced in
Appendix~\ref{app:example}.

Throughout we impose the \emph{finite-support} condition: only finitely many cells
are initially black. The empirical universality of the highway is the subject of a
well-known open problem.

\begin{conjecture}[Highway conjecture]\label{conj:main}
For every finite initial black set and every initial pose, there is a time after which
the ant's future turn trace and path agree with the standard period-$104$ highway up
to lattice translation, quarter-turn rotation, and cyclic phase shift. Finite cells
outside the future path may remain as stationary debris; equality of the entire
colouring is not asserted.
\end{conjecture}

\subsection{Prior art}
The rigorous facts about the classical ant are few. Every trajectory is
\emph{unbounded} (Bunimovich--Troubetzkoy \cite{BunimovichTroubetzkoy1992}; the
underlying checkerboard/parity structure is already in Gale--Propp--Sutherland--Troubetzkoy
\cite{GaleProppSutnerTroubetzkoy1995}). The ant is computationally universal, and
associated prediction problems are undecidable, on suitable infinite backgrounds
\cite{GajardoMoreiraGoles2002}; we draw no inference from this about
Conjecture~\ref{conj:main}, which concerns finite configurations, beyond the evident
point that no finite cutoff exhausts an infinite range. Gajardo
\cite{Gajardo2003symbolic} gives an algebraic characterisation of the forbidden
factors of the ant's trace subshift. Recent work on \emph{generalised} ants exhibits
rules with infinitely many highways \cite{GajardoLutfallaRao2024} and rules with
non-highway emergent behaviour \cite{Lutfalla2025sideways}. Quantitative confinement
bounds for the classical ant are studied in \cite{Etse2026confined}. The present
paper concerns a different, conditional question: what restrictions follow once a
classical finite-support orbit is assumed to have a periodic translating tail?

\subsection{Results}
A \emph{repeating trace} is an infinite turn word $w^\omega$ realised from a
finite-support colouring (Section~\ref{sec:criterion}), where the finite word $w$
resets the heading; its \emph{drift}
$\drift=(a,b)$ is the per-period displacement and its \emph{growth} $\growth$ the
per-period change in the number of black cells. The main results are:

\begin{enumerate}[leftmargin=2.4em,label=\textbf{(\arabic*)}]
\item \textbf{Even-winding theorem} (Theorem~\ref{thm:evenwinding}): no
finite-support clean translator with nonzero drift exists. Hence every
finite-support periodic highway has growth $\growth\in4\Z_{>0}$
(Corollary~\ref{cor:growth4}). Bunimovich--Troubetzkoy excludes only \emph{bounded}
repeatable orbits; a zero-growth glider is unbounded and is not covered by their
theorem. (The divisibility by $4$ alone is the
elementary parity Lemma~\ref{lem:invariant}.)

\item \textbf{Signed mod-four wake-residue identity.}
Theorem~\ref{thm:residue}; Corollary~\ref{cor:P16} gives the lower bound
\[
\growth\ge2\max(|a|,|b|).
\]

\item \textbf{Tait-graph conjugacy} (Theorem~\ref{thm:conjugacy}) and the cycle-rank
surgery identity $E+\Delta C=2\beta$ (Theorem~\ref{thm:cyclerank}), built on the
checkerboard/HV partition of \cite{GaleProppSutnerTroubetzkoy1995}.

\item \textbf{Collision-chain parity} (Theorem~\ref{thm:collision}), a two-endpoint
$\F_2$ boundary identity for colour defects.

\item \textbf{Transverse rigidity and width-two/width-four exclusions}
(Theorems~\ref{thm:extremal}, \ref{thm:width2}, and~\ref{thm:width4}). For diagonal drift the two
extremal level lines of a periodic highway are horizontal, every turn on them is
$\turnR$, and each cell reached on them is visited exactly once by the periodic tail
and is left black; the transverse width is even. No such highway has
transverse widths two or four; hence its width is at least six. Unlike~(6), these
statements are not bounded period searches.  The width-four theorem is
computer-assisted: two independent exact local enumerators certify a finite
twelve-edge blank-column graph, and Lean certifies that every listed edge lowers
rank.  The local enumeration's completeness is not yet formalised in Lean.  The
proofs are independent of~(2).

\item \textbf{Small-period exclusion} (Theorem~\ref{thm:periodexcl}): a rank-audited
exhaustive search excludes every finite-support periodic highway of nonzero drift and
period at most $48$. The search assumes no consequence of the residue theorem and
applies the criterion to every positive-growth nonzero-drift leaf, so the exclusion
does not depend on Section~\ref{sec:residue}; faster variants that do assume it agree.
The certifying search is deliberately separated from the faster conditional
cross-checks described in Appendix~\ref{app:repro}.
\end{enumerate}

Section~\ref{sec:criterion} also gives a decidable periodic-realisability criterion
with an explicit finite seed, strengthening the forbidden-word program of
\cite{Gajardo2003symbolic} from local factor-admissibility to global periodic
finite-support realisability. The
elementary heading/black-count invariant (Lemma~\ref{lem:invariant}) is folklore,
implicit in \cite{BunimovichTroubetzkoy1992,GaleProppSutnerTroubetzkoy1995}.

\subsection{Scope}
Every theorem below is a \emph{necessary condition} on a finite-support periodic
highway, or an exact finite computation. None asserts that any orbit converges, and
therefore none resolves Conjecture~\ref{conj:main}. What we prove is that any
finite-support periodic highway must build a positive wake divisible by four and must
satisfy the residue identities and the density bound $\growth\ge2\|\drift\|_\infty$,
and that no finite-support periodic highway has period at most $48$ (equivalently, any
such highway has period at least $50$). Two logically separate gaps remain: an
arbitrary finite-support orbit has not been shown to enter a translating periodic
regime, and positive-growth finite-support periodic traces have not been classified.
Section~\ref{sec:gap} states both problems explicitly.

%=======================================================================
\section{Setup and elementary facts}\label{sec:setup}
%=======================================================================

A \emph{state} is a triple $s=(B,p,q)$ with $B\subset\Z^2$ finite (the black set),
$p\in\Z^2$ (the ant position) and $q\in\Z/4\Z$ the heading, numbered $0=N$, $1=E$,
$2=S$, $3=W$, with unit vectors
\[
u_0=(0,1),\quad u_1=(1,0),\quad u_2=(0,-1),\quad u_3=(-1,0).
\]
The \emph{turn} at $s$ is $\turnR$ (right) with $\varepsilon=+1$ if $p\notin B$, and
$\turnL$ (left) with $\varepsilon=-1$ if $p\in B$. One step of the dynamics is
\begin{equation}\label{eq:step}
s=(B,p,q)\ \longmapsto\ \bigl(B\,\triangle\,\{p\},\ p+u_{q+\varepsilon},\ q+\varepsilon\bigr),
\end{equation}
where $\triangle$ denotes symmetric difference and indices are read modulo $4$. The
\emph{trace} of an orbit is the infinite word of turns encountered. Since
\eqref{eq:step} is invertible (step backward one cell, recolour, undo the turn), the
dynamics is reversible and orbits never merge. Write $b_t=|B_t|$ and let
$\che(x,y)=(-1)^{x+y}$ be the checkerboard sign.

\begin{definition}[Periodic terminology]\label{def:periodic}
Let $w=w_0\cdots w_{N-1}$ be a nonempty turn word and follow its induced lattice
path from a specified initial heading. The word is \emph{heading-resetting} if the
heading after $N$ turns equals the initial heading. In that case its \emph{drift} is
$p_N-p_0$, and its \emph{growth} is
$\#\{i:w_i=\turnR\}-\#\{i:w_i=\turnL\}$. An orbit has an \emph{eventually periodic
turn trace} if, from some time $t_0$, its turns are $w^\omega$. If $w$ is
heading-resetting with nonzero drift, we call that tail a \emph{periodic highway};
its \emph{least trace period} is the smallest possible $N$. Highways are compared up
to lattice translation, quarter-turn rotation, and cyclic phase shift of $w$.
\end{definition}

\begin{proposition}[Checkerboard/HV partition {\cite{GaleProppSutnerTroubetzkoy1995}}]
\label{prop:hv}
The quantity $\bigl(q_t-(p_{t,x}+p_{t,y})\bigr)\bmod 2$ is constant along every
orbit. Consequently each cell is entered by the ant always horizontally or always
vertically. Calling a cell \emph{vertical} (resp.\ \emph{horizontal}) accordingly,
at a vertical cell the two vertical incident edges are the only admissible arrival
edges and the two horizontal ones the only departure edges, and vice versa; every
admissible directed lattice edge thus carries a fixed orientation.
\end{proposition}

\begin{proof}
Each step changes the heading by an odd quarter-turn, so $q_t\bmod2$ alternates;
the displacement $u_{q_{t+1}}$ has odd $\ell^1$-length, so $(p_{t,x}+p_{t,y})\bmod2$
alternates as well. Their difference is therefore invariant. Fixing a cell $z$: at
every visit the checkerboard parity of $z$ is the same, so $q_t\bmod2$ is the same,
i.e.\ the arrival axis is fixed. The edge orientations follow because an arrival
edge of one endpoint is a departure edge of the other.
\end{proof}

\begin{proposition}[Alternating visits]\label{prop:alt}
At any fixed cell the encountered turns alternate strictly between $\turnR$ and
$\turnL$. All but finitely many cells begin with $\turnR$.
\end{proposition}

\begin{proof}
Each visit flips the cell's colour and the turn is a function of the colour, so the
turn alternates. A cell white at its first visit begins with $\turnR$; only the
finitely many initially black cells can begin with $\turnL$.
\end{proof}

\begin{theorem}[Unboundedness {\cite{BunimovichTroubetzkoy1992}}]\label{thm:unbounded}
Every finite-support orbit visits infinitely many cells.
\end{theorem}

\begin{proof}[Proof sketch]
Were the visited set finite, determinism and reversibility would make the orbit
periodic in the full state. Choose a visited cell $z$ maximising $x+y$; its north and
east neighbours are unvisited. If $z$ is horizontal it can only be entered from the
west. During a full state period its colour returns, so its alternating visits include
both a white/$\turnR$ and a black/$\turnL$ visit; these two turns leave respectively
south and north, forcing a visit to the unvisited north neighbour. If $z$ is vertical
it can only be entered from the south, and the two colours force departures east and
west, including the unvisited east neighbour. Either case is a contradiction.
\end{proof}

\begin{lemma}[Heading/black-count invariant; folklore]\label{lem:invariant}
For every $t$,
\begin{equation}\label{eq:invariant}
q_t-b_t\equiv q_0-b_0\pmod 4,\qquad p_{t+1}-p_t=u_{\,q_0-b_0+b_{t+1}} .
\end{equation}
A heading-resetting period changes $b$ by a multiple of $4$.
\end{lemma}

\begin{proof}
A white/$\turnR$ step increments both $q$ and $b$ by $1$; a black/$\turnL$ step
decrements both. Hence $q_t-b_t$ is constant. Since the move follows the turn,
$q_{t+1}\equiv q_0-b_0+b_{t+1}$, giving the displacement formula. If $q_N=q_0$ then
$b_N-b_0\equiv 0\pmod4$.
\end{proof}

%=======================================================================
\section{A constructive periodic-realisability criterion}\label{sec:criterion}
%=======================================================================

We first record the exact test that turns ``is $w^\omega$ a highway?'' into a finite
computation. Fix an initial heading $q_0$ and a proposed turn word
$w=w_0\cdots w_{N-1}\in\{\turnR,\turnL\}^N$. Compute the induced path
$p_0,\dots,p_N$ from $w$ (Proposition~\ref{prop:hv} makes each step well defined).
Suppose $q_N=q_0$ and $\drift=p_N-p_0\neq0$. Define an equivalence on phases by
\[
i\sim j\iff p_i-p_j\in\Z\drift,
\]
and in each class $I$ write $p_i=b_I+a_i\drift$ for a fixed representative $b_I$ and
integers $a_i$. For $n\ge\min_I a_i$, let $S_{I,n}$ be the subword of all $w_i$
($i\in I$) with $a_i\le n$, listed in increasing order of the key $(n-a_i,\,i)$.

\begin{theorem}[Periodic-realisability criterion]\label{thm:criterion}
The infinite word $w^\omega$ is the trace of Langton's ant from some finite-support
colouring if and only if, for every class $I$:
\begin{enumerate}[label=\textup{(\roman*)}]
\item every $S_{I,n}$ with $\min_I a_i\le n\le\max_I a_i$ alternates strictly between
$\turnR$ and $\turnL$; and
\item the stabilised word $S_{I,\max_I a_i}$ begins with $\turnR$.
\end{enumerate}
When these hold, an explicit finite black seed producing $w^\omega$ is
\begin{equation}\label{eq:seed}
B=\bigl\{\,b_I+n\drift\ :\ \textstyle\min_I a_i\le n<\max_I a_i\ \text{and}\ S_{I,n}\ \text{begins with}\ \turnL\,\bigr\},
\end{equation}
the union over all classes $I$; the bound $n<\max_I a_i$ together with
$n\ge\min_I a_i$ makes $B$ finite and $S_{I,n}$ well defined.
\end{theorem}

\begin{proof}
The occurrence of phase $i\in I$ in period $k\ge0$ sits at
$p_i+k\drift=b_I+(a_i+k)\drift$, which equals $b_I+n\drift$ exactly when $k=n-a_i\ge0$;
chronological order among these occurrences is the lexicographic order of $(n-a_i,i)$.
Thus $S_{I,n}$ is precisely the chronological turn sequence at the physical cell
$b_I+n\drift$ under $w^\omega$.

\emph{Necessity.} Each visit flips the cell, so its turn sequence alternates,
proving (i). For $n\ge\max_I a_i$ all phases of $I$ are present and the ordering no
longer depends on $n$: the sequence has stabilised, and the infinitely many cells
$b_I+n\drift$ ($n\ge\max_I a_i$) share this first turn. By Proposition~\ref{prop:alt}
only finitely many cells begin with $\turnL$, so the stable first turn is $\turnR$,
proving (ii).

\emph{Sufficiency.} Colour exactly the set \eqref{eq:seed} black. At each boundary
cell $b_I+n\drift$ with $n<\max_I a_i$ the chosen colour reproduces the first symbol
of $S_{I,n}$, and strict alternation supplies every subsequent symbol because the
cell flips at each visit. At every stabilised cell ($n\ge\max_I a_i$) condition (ii)
makes the first encountered turn $\turnR$, matching its white initial colour.
Induction on chronological time shows the ant follows $w^\omega$ forever; since
$q_N=q_0$ and the position advances by $\drift\neq0$ each period, this is a highway.
\end{proof}

\begin{remark}\label{rem:onesort}
There is an equivalent single-sequence test: within a class, sort all entries by
decreasing $a_i$, breaking ties by increasing $i$; every $S_{I,n}$ is a suffix of the
result obtained by deleting whole high-level blocks. It therefore suffices to check
that this one stabilised word alternates and begins with $\turnR$, at the cost of one
sort per class. To index the classes of $\Z^2/\Z\drift$ exactly, write
$g=\gcd(a,b)$ and $\drift=g\,(a',b')$ with $\gcd(a',b')=1$, and choose integers $r,s$
with $r a'+s b'=1$ (B\'ezout). Then
\[
\bigl(\,b'x-a'y,\ (r x+s y)\bmod g\,\bigr)
\]
is a \emph{complete} invariant of the coset of $(x,y)$: the matrix
$\left(\begin{smallmatrix} b'&-a'\\ r&s\end{smallmatrix}\right)$ has determinant
$b's+a'r=1$, so it is unimodular, and in its coordinates translation by
$\drift=g(a',b')$ becomes translation by $(0,g)$, leaving the first coordinate and the
second coordinate modulo $g$ fixed. (The naive residue $a'x+b'y\bmod g$ is
\emph{not} complete: for $\drift=(2,2)$ the cells $(0,0)$ and $(1,1)$ share both
$b'x-a'y=0$ and $a'x+b'y\equiv0\pmod2$ yet lie in different cosets, since
$(1,1)\notin\Z(2,2)$; the B\'ezout coordinate $rx+sy$ separates them.) This is the
checker used in all computations
of Section~\ref{sec:computation}; it strengthens the
forbidden-word characterisation of the ant's trace subshift
\cite{Gajardo2003symbolic} to a decidable test for global periodic finite-support
realisability with a constructive seed. A worked instance (the standard highway) is
given in Appendix~\ref{app:example}.
\end{remark}

%=======================================================================
\section{The Tait-graph conjugacy}\label{sec:tait}
%=======================================================================

We recast the dynamics on a medial (Tait) graph, exposing topology invisible in the
raw colouring. Normalise the HV partition of Proposition~\ref{prop:hv} so that
even cells receive vertical arrivals. In the \emph{frozen} routing (the permutation
of directed states obtained by ignoring colour flips) the admissible directed arcs
form clockwise directed four-cycles around one checkerboard class of unit plaquettes.
Label a plaquette by its lower-left corner and take these plaquettes as the vertices
of a rotated square lattice. Each ant cell $z=(x,y)$ lies at the crossing of exactly
two white four-cycles and corresponds to the \emph{Tait edge}
\begin{equation}\label{eq:tait}
e_z=\begin{cases}\{(x,y-1),(x-1,y)\}, & x+y\ \text{even},\\[2pt]
\{(x-1,y-1),(x,y)\}, & x+y\ \text{odd}.\end{cases}
\end{equation}
Let $H_B=\{e_z:z\in B\}$ be the finite edge subgraph induced by the black cells.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=1.05,line width=0.9pt]
  % --- crossing at z, with the four incident arcs labelled 1..4
  \begin{scope}
    \draw (-0.8,-0.8)--(0.8,0.8);
    \draw (-0.8,0.8)--(0.8,-0.8);
    \filldraw (0,0) circle (1.6pt);
    \node[font=\scriptsize] at (-1.02,-1.02) {$1$};
    \node[font=\scriptsize] at (-1.02, 1.02) {$2$};
    \node[font=\scriptsize] at ( 1.02, 1.02) {$3$};
    \node[font=\scriptsize] at ( 1.02,-1.02) {$4$};
    \node[font=\scriptsize] at (0,-0.42) {$z$};
    \node at (0,-1.5) {\footnotesize crossing at $z$};
  \end{scope}
  % --- white smoothing: joins 1-2 and 3-4 (e_z absent)
  \begin{scope}[xshift=3.9cm]
    \draw (-0.8,-0.8) .. controls (-0.15,-0.35) and (-0.15,0.35) .. (-0.8,0.8);
    \draw (0.8,-0.8)  .. controls (0.15,-0.35)  and (0.15,0.35)  .. (0.8,0.8);
    \node[font=\scriptsize] at (-1.02,-1.02) {$1$};
    \node[font=\scriptsize] at (-1.02, 1.02) {$2$};
    \node[font=\scriptsize] at ( 1.02, 1.02) {$3$};
    \node[font=\scriptsize] at ( 1.02,-1.02) {$4$};
    \node at (0,-1.5) {\footnotesize white: $1{-}2$, $3{-}4$; $e_z\notin H_B$};
  \end{scope}
  % --- black smoothing: joins 1-4 and 2-3; Tait edge e_z present (dashed)
  \begin{scope}[xshift=7.8cm]
    \draw (-0.8,-0.8) .. controls (-0.35,-0.15) and (0.35,-0.15) .. (0.8,-0.8);
    \draw (-0.8,0.8)  .. controls (-0.35,0.15)  and (0.35,0.15)  .. (0.8,0.8);
    \draw[dashed] (0,-0.42)--(0,0.42);
    \node[font=\scriptsize] at (0.34,0) {$e_z$};
    \node[font=\scriptsize] at (-1.02,-1.02) {$1$};
    \node[font=\scriptsize] at (-1.02, 1.02) {$2$};
    \node[font=\scriptsize] at ( 1.02, 1.02) {$3$};
    \node[font=\scriptsize] at ( 1.02,-1.02) {$4$};
    \node at (0,-1.5) {\footnotesize black: $1{-}4$, $2{-}3$; $e_z\in H_B$};
  \end{scope}
\end{tikzpicture}
\caption{The two smoothings of the frozen crossing at an ant cell $z$. The four
incident frozen arcs are labelled $1,2,3,4$. Colour selects which of the two ways they
reconnect: white joins $1{-}2$ and $3{-}4$, black joins $1{-}4$ and $2{-}3$. The black
smoothing is exactly the occupied Tait edge $e_z$ (dashed).}
\label{fig:smoothing}
\end{figure}

\begin{theorem}[Boundary-loop conjugacy]\label{thm:conjugacy}
White and black at $z$ are the two smoothings of the directed crossing at the Tait
edge $e_z$ \textup{(Figure~\ref{fig:smoothing})}. Consequently:
\begin{enumerate}[label=\textup{(\alph*)}]
\item the frozen routing loops are exactly the boundary components of a regular
neighbourhood of $H_B$, together with the untouched four-cycles;
\item the ant advances along its current frozen boundary loop and then toggles the
edge $e_z$; and
\item each step either merges two frozen loops into one or splits one into two.
\end{enumerate}
\end{theorem}

\begin{proof}
Frozen (colour-ignoring) routing is the permutation $\rho$ of directed edges of the
ant lattice sending each directed edge to the next one the ant would follow under a
fixed local turn rule; since a white cell turns right and a black cell turns left, at
each ant cell $z$ the local routing is one of the two ways of connecting the four
directed edge-ends incident to $z$ that avoids a $U$-turn. These are precisely the two
\emph{smoothings} of the $4$-valent vertex $z$ of the medial graph, and the Tait edge
$e_z$ of \eqref{eq:tait} records which diagonal pair of plaquette-corners the crossing
joins; white is one smoothing and black the other, which is (a) at the level of a
single crossing.

Now freeze a colouring $B$. Away from $H_B$ every plaquette crossing is white, and the
white smoothings assemble into the boundary circuits of a regular neighbourhood $\mathcal{N}$
of the edge subgraph $H_B$: each occupied edge $e_z$ is thickened to a band, each Tait
vertex to a disk, and the frozen routing runs along $\partial\mathcal{N}$, with the untouched
white plaquettes contributing the trivial four-cycles. (This is the standard medial/Tait
correspondence between a plane graph and the boundary of a ribbon neighbourhood.) This
gives (a) globally: the frozen loops are the components of $\partial\mathcal{N}$ plus the
untouched four-cycles. Statement (b) is immediate: the live ant sits on the current
boundary loop and, on stepping off $z$, toggles the colour of $z$, i.e.\ replaces the
smoothing at that single crossing---adding or deleting the band $e_z$ in $\mathcal{N}$.

For (c), toggling $e_z$ is the addition or deletion of one band at a $4$-valent
vertex, which is a single Dehn-twist-free band surgery on $\partial\mathcal{N}$. Locally it
reconnects the two boundary arcs passing the vertex: if the two arcs belong to the same
boundary component the surgery splits it into two, and if they belong to different
components it merges them into one. There are no other cases, giving (c).
\end{proof}

Let $E=|B|$ be the number of occupied Tait edges, $V$ the number of incident Tait
vertices, $k$ the number of nonempty components, and $\beta=E-V+k$ the cycle rank.

\begin{theorem}[Cycle-rank surgery identity]\label{thm:cyclerank}
A regular neighbourhood of $H_B$ has $k+\beta=E-V+2k$ boundary components. Writing
$\Delta C$ for the change in loop count relative to the $V$ trivial plaquette loops,
\begin{equation}\label{eq:cyclerank}
\Delta C=E-2V+2k,\qquad E+\Delta C=2\beta .
\end{equation}
Adding a bridge merges two loops and fixes $\beta$; adding a cycle edge splits a loop
and raises $\beta$ by one; removing a bridge splits a loop and fixes $\beta$;
removing a non-bridge merges two loops and lowers $\beta$ by one.
\end{theorem}

\begin{proof}
A regular neighbourhood of a plane graph with $V$ vertices, $E$ edges, and $k$
components is a compact planar surface with $k$ genus-zero pieces and Euler
characteristic $V-E$. If it has $b$ boundary circles, then $V-E=2k-b$, so
$b=E-V+2k=k+\beta$, where $\beta=E-V+k$ is the first Betti number.
Formula~\eqref{eq:cyclerank} is the difference from the $V$ trivial plaquette loops.
The four surgery cases are the standard effect on the boundary of adding or deleting
a bridge (Betti number unchanged; boundary count changes by $\mp1$) versus a cycle
edge (Betti number changes by $\pm1$; opposite merge/split), read off from
Theorem~\ref{thm:conjugacy}(c).
\end{proof}

\begin{proposition}[Advanced records are bridges]\label{prop:record}
Once a new bounding-box record lies strictly beyond the bounding box of the finite
initial support, its first visit is white, and a new east, north, west, or south
record causes the ant to turn south, east, north, or west, respectively. In the
Tait graph the corresponding edge has a previously untouched endpoint, hence is a
bridge. Therefore sufficiently advanced record creation never raises $\beta$; any
unbounded production of cycle rank occurs strictly inside already-explored territory.
\end{proposition}

\begin{proposition}[No monotone affine invariant]\label{rem:nomonotone}
No nonconstant affine combination of $(E,V,k,\beta)$ is monotone along the blank
orbit. In particular, each of $E$, $V$, $k$, and $\beta$ separately fails to be
monotone. Thus no affine combination of these four graph statistics can serve as a
monotone or Lyapunov potential along even the blank orbit.
\end{proposition}

\begin{proof}
Since $\beta=E-V+k$ identically,
$aE+bV+ck+d\beta+e=(a+d)E+(b-d)V+(c+d)k+e$, so the affine combinations of
$(E,V,k,\beta)$ are exactly those of $(E,V,k)$, and such a combination is nonconstant
precisely when $\lambda:=(a+d,\,b-d,\,c+d)\in\mathbb{R}^3$ is nonzero.

One step flips one cell, hence inserts or deletes exactly one Tait edge
$e_z=\{u,v\}$. Insertion admits exactly four cases, according to how many endpoints of
$e_z$ already meet an edge of $H_B$ and, when both do, whether they lie in a common
component; deletion reverses each. This is a complete case list, not a sample:
\begin{center}
\begin{tabular}{@{}lcc@{}}
\toprule
insertion of $e_z=\{u,v\}$ & $(\Delta E,\Delta V,\Delta k)$ & $\Delta\beta$\\
\midrule
$u,v$ both new & $(1,2,1)$ & $0$\\
exactly one of $u,v$ new & $(1,1,0)$ & $0$\\
$u,v$ both old, distinct components & $(1,0,-1)$ & $0$\\
$u,v$ both old, same component & $(1,0,0)$ & $1$\\
\bottomrule
\end{tabular}
\end{center}
Hence every increment lies in
\begin{equation}\label{eq:evkdeltas}
S_0=\pm\bigl\{(1,2,1),\,(1,1,0),\,(1,0,-1),\,(1,0,0)\bigr\},
\end{equation}
and, incidentally, $\beta$ rises exactly when the inserted edge closes a cycle and is
unchanged otherwise. All eight members of $S_0$ are realised on the blank orbit---every
state of which is compact---within the first $12{,}000$ steps, so the realised
increment set is exactly $S_0$.

Now $S_0$ is centrally symmetric by construction, since deleting an edge negates the
increment of inserting it, and $S_0$ contains the linearly independent
$(1,1,0),(1,0,-1),(1,0,0)$, so $\operatorname{span}S_0=\mathbb{R}^3$. Let
$\lambda\neq0$. Spanning gives $v\in S_0$ with $\lambda\cdot v\neq0$; replacing $v$ by
$-v\in S_0$ if necessary, $\lambda\cdot v>0$, while $\lambda\cdot(-v)<0$ and $-v$ is
realised as well. Hence $\lambda$ strictly increases at one step of the orbit and
strictly decreases at another, so it is monotone in neither direction. Equivalently,
$0$ lies in the interior of $\operatorname{conv}S_0$. The realisability witnesses are
reproduced by \texttt{verify\_nomonotone.py} (Appendix~\ref{app:repro}).
\end{proof}

%=======================================================================
\section{Additive charges and clean translators}\label{sec:charges}
%=======================================================================

\begin{definition}\label{def:translator}
A \emph{clean finite translator with nonzero drift} is a finite state for which, for
some $N>0$ and $\drift\in\Z^2\setminus\{0\}$,
\[
B_N=B_0+\drift,\qquad p_N=p_0+\drift,\qquad q_N=q_0 .
\]
The word \emph{clean} means the entire black set translates: no stationary residue
and no newly emitted wake. This is not the ordinary highway, which grows a wake.
\end{definition}

\subsection{Laurent charges and axis alignment}
Work in the Laurent polynomial ring $\F_2[X^{\pm1},Y^{\pm1}]$, an integral domain.
Let $V_t=1$ if $q_t$ is vertical and $H_t=1-V_t$, and set
\[
A_t=\sum_{(x,y)\in B_t}X^x+V_t\,X^{x_t},\qquad
D_t=\sum_{(x,y)\in B_t}Y^y+H_t\,Y^{y_t}.
\]

\begin{proposition}\label{prop:laurent}
$A_t$ and $D_t$ are conserved. Hence a clean translator with $\drift=(a,b)$ satisfies
$A=X^aA$ and $D=Y^bD$; since $A(1)=|B|+V$ and $D(1)=|B|+H$ cannot both vanish in
$\F_2$ (because $V+H=1$), the drift coordinates $a$ and $b$ are not both nonzero.
Every clean translator is axis-aligned.
\end{proposition}

\begin{proof}
Conservation of $A_t$: on a step from a vertical cell, the toggle at column $x_t$
cancels the disappearing vertical correction $X^{x_t}$; on a step from a horizontal
cell, the post-turn motion is vertical, so $x_{t+1}=x_t$ and the toggle cancels the
newly appearing correction. Translation covariance then gives $A=X^aA$. In the
integral domain $\F_2[X^{\pm1},Y^{\pm1}]$, $a\neq0$ forces $A=0$ and $b\neq0$ forces
$D=0$; but $A(1)+D(1)=2|B|+1=1\neq0$.
\end{proof}

\begin{proposition}[Additive-charge cocycle]\label{prop:cocycle}
Let $A$ be an abelian group and $w:\Z^2\to A$. If there exist heading corrections
$F_q:\Z^2\to A$ (indexed by $q\in\Z/4\Z$) such that $\sum_{z\in B}w(z)+F_q(p)$ is
conserved under
\begin{equation}\label{eq:pot}
F_{q+1}(z+u_{q+1})-F_q(z)=-w(z)\ \ (\turnR),\qquad
F_{q-1}(z+u_{q-1})-F_q(z)=+w(z)\ \ (\turnL),
\end{equation}
then $w$ satisfies both the four-corner condition
\begin{equation}\label{eq:cocycle}
w(x{+}1,y{+}1)+w(x{+}1,y)+w(x,y{+}1)+w(x,y)=0\quad\text{in }A
\end{equation}
and $4w=0$. Conversely, if $A=\Z/4\Z$ \emph{(}so $4w=0$ holds automatically\emph{)}
and $w$ satisfies \eqref{eq:cocycle}, then such $F_q$ exist. Over $\F_2$ the solutions
of \eqref{eq:cocycle} are exactly $w(x,y)=\alpha(x)+\beta(y)$.
\end{proposition}

\begin{proof}
\emph{Necessity.} Eliminating the four $F_q$ from \eqref{eq:pot} around a unit
plaquette gives the four-corner condition \eqref{eq:cocycle}. For $4w=0$, combine the
four relations at a single cell $z$: the $\turnR$-at-$S$ and $\turnL$-at-$N$ relations
both express $F_W(z+u_W)$, forcing $F_S(z)=F_N(z)+2w(z)$, while the $\turnL$-at-$S$ and
$\turnR$-at-$N$ relations force $F_S(z)=F_N(z)-2w(z)$; hence $4w(z)=0$. Thus
$w(x,y)=(-1)^{x+y}$ over $A=\Z$ satisfies \eqref{eq:cocycle} but admits no such $F_q$,
since $4w\neq0$.

\emph{Sufficiency for $A=\Z/4\Z$.} Composing pairs of relations in \eqref{eq:pot}
expresses $F_N$ by its diagonal increments
\begin{equation}\label{eq:Fdiag}
\begin{aligned}
F_N(x{+}1,y{+}1)-F_N(x,y)&=-w(x,y)+w(x{+}1,y),\\
F_N(x{+}1,y{-}1)-F_N(x,y)&=w(x,y{-}1)-w(x{+}1,y{-}1).
\end{aligned}
\end{equation}
which define $F_N$ on each of the two $x{+}y$-parity sublattices by integration from a
basepoint. Modulo four, the circulation of \eqref{eq:Fdiag} around the elementary
diamond $(x,y)\to(x{+}1,y{+}1)\to(x{+}2,y)\to(x{+}1,y{-}1)\to(x,y)$ equals the negative
of the sum of the four-corner expressions \eqref{eq:cocycle} based at $(x,y{-}1)$ and
$(x{+}1,y{-}1)$ (the two differ by $4\,[w(x{+}1,y)+w(x{+}1,y{-}1)]\equiv0$); hence it
vanishes by \eqref{eq:cocycle}, so the integration is path-independent and $F_N$ is
well defined. Now set $F_S=F_N-2w$ and define $F_E,F_W$ from $F_N$ by \eqref{eq:pot}.
Substituting these definitions back into the four $\turnR/\turnL$ relations verifies
each case in turn; the only possible ambiguity is between the two ways of reaching
$F_S$---via the $\turnR$-at-$S$/$\turnL$-at-$N$ pair, which gives $F_N+2w$, and via the
$\turnL$-at-$S$/$\turnR$-at-$N$ pair, which gives $F_N-2w$---and their discrepancy is
exactly $4w$, which vanishes. Hence all four $F_q$ are consistent. Over $\F_2$ the
solution space of \eqref{eq:cocycle} is the separable weights $\alpha(x)+\beta(y)$.
\end{proof}

\subsection{The strand decomposition}
Let $\growth=\#\turnR-\#\turnL$ over one period. Lemma~\ref{lem:invariant} gives
$\growth\equiv0\pmod4$, and $\growth\ge0$ because $\growth$ equals the number of
\emph{odd} stabilised translation classes of Theorem~\ref{thm:criterion} (a class is
odd iff its stabilised word has odd length, contributing one net black cell per
period). We record the endpoint structure used in Section~\ref{sec:residue}.

\begin{proposition}[Canonical endpoint decomposition]\label{prop:decomp}
The one-period signed toggle pattern $D:\Z^2\to\Z$ (with $D(z)=\#\{\turnR\text{ at }z\}-\#\{\turnL\text{ at }z\}$ over one period) decomposes as
\begin{equation}\label{eq:decomp}
D=S+(T_\drift-1)A,\qquad S=\sum_{i=1}^{\growth}\delta_{c_i},
\end{equation}
where $(T_\drift A)(z)=A(z-\drift)$, $A$ is a finitely supported \emph{moving widget},
and $c_1,\dots,c_\growth\in\Z^2$ are the bases of the $\growth$ growing strands, one
per odd stabilised class. For the standard word $\growth=12$ and $|B_n|=11+12n$ at
period boundaries.
\end{proposition}

\begin{proof}
$D$ is finitely supported (only finitely many cells are visited in one period).
Partition the support into translation classes under $z\mapsto z+\drift$; there are
finitely many nonempty classes. Fix a class with base point $c$ and write
$f(n)=D(c+n\drift)\in\Z$, a finitely supported function $\Z\to\Z$. Put $s=\sum_n f(n)$
and $h=f-s\,\delta_0$, so $\sum_n h(n)=0$. With the shift $(Ta)(n)=a(n-1)$, set
\[
a(n)=\sum_{m>n}h(m).
\]
Then $a(n-1)-a(n)=h(n)$, i.e.\ $(T-1)a=h$, so
\[
f=s\,\delta_0+(T-1)a,
\]
and $a$ is finitely supported: $a(n)=0$ for $n$ beyond the support of $f$ on the
right, and $a(n)=\sum_m h(m)=0$ for $n$ far to the left (this is exactly where
$\sum h=0$ is needed). Summing over all classes gives \eqref{eq:decomp} with $A$ the
(finite) sum of the class antiderivatives and $S=\sum_{\text{classes}}s_c\,\delta_{c}$. It remains to
identify $\sum_n f$ for a class. Over one period the cells of a class are visited in
level order, and the alternating-visit rule (Proposition~\ref{prop:alt}) makes the
per-class toggle total equal to $+1$ when the class's stabilised word has odd length
(an ``odd class'') and $0$ when it has even length; thus $\sum_n f\in\{0,1\}$ and
$S$ is a sum of $\growth$ unit masses, one per odd class, where
$\growth=\#\{\text{odd classes}\}=\#\turnR-\#\turnL$ (the last equality because both
count the net black cells created per period). The standard-word figures follow from
$|B_0|=11$ and $|B_{n+1}|-|B_n|=\growth=12$.
\end{proof}

\begin{corollary}\label{cor:sep}
If $\growth=0$ then a finite stationary set can be deleted to leave a clean finite
translator with drift $\drift$.
\end{corollary}

\begin{proof}
Here $S=0$, so $D=(T_\drift-1)A$. Let $b_n:\Z^2\to\{0,1\}$ be the indicator of $B_n$
at the $n$-th period boundary. Since each cell is toggled a net $D(z)\in\{-1,0,+1\}$
per period (Proposition~\ref{prop:alt}) and the period-$k$ toggle pattern is
$T_{k\drift}D$ by translation covariance,
\[
b_n=b_0+\sum_{k=0}^{n-1}T_{k\drift}D
   =b_0+\sum_{k=0}^{n-1}T_{k\drift}(T_\drift-1)A
   =b_0+(T_{n\drift}-1)A=R+T_{n\drift}A,\qquad R:=b_0-A .
\]
Both $R$ and $A$ are finitely supported, so for $n$ large the supports of $R$ and
$T_{n\drift}A$ are disjoint; since $b_n$ is $\{0,1\}$-valued, $R$ and $A$ are then
$\{0,1\}$-valued and $B_n=R\sqcup(A+n\drift)$. Because the trace is periodic with
nonzero drift, each cell is visited finitely often, so at a sufficiently late period
boundary the ant never again visits the finite set $R$. Deleting $R$ there changes no
future turn and leaves a state whose black set is $A+n\drift$, translating exactly by
$\drift$ each period: a clean finite translator.
\end{proof}

Sections~\ref{sec:evenwinding} and~\ref{sec:residue} prove, by two independent
routes, that no such clean translator exists; hence $\growth\neq0$.

%=======================================================================
\section{The even-winding theorem}\label{sec:evenwinding}
%=======================================================================

This section proves that no clean translator exists, via a cylinder homology
argument. The one step that was only sketched in earlier accounts---the local
parity identity relating the turn count at a cell to its horizontal edge
parity---is proved in full as Lemma~\ref{lem:localparity}.

\begin{theorem}[Even-winding theorem]\label{thm:evenwinding}
No finite-support clean translator with nonzero drift exists.
\end{theorem}

For the remainder of the section fix a hypothetical clean translator. Its drift is
axis-aligned by Proposition~\ref{prop:laurent}. A clean translator has $\growth=0$, so
its period $N=\#\turnR+\#\turnL=2\#\turnR$ is even; since each step changes $x+y$
modulo $2$, the total displacement satisfies $a+b\equiv N\equiv0\pmod2$, so the
nonzero drift coordinate is even. After a quarter-turn rotation write $\drift=(L,0)$
with $L>0$ even---all that the argument below requires. Writing $B_j,p_j$ for the black
set and ant position at phase $j$ of one period, reversibility and translation
covariance give, for all $n\in\Z$,
\[
B_{nN+j}=B_j+n\drift,\qquad p_{nN+j}=p_j+n\drift .
\]
Quotient the bi-infinite trace by translation through $\drift$ to obtain a closed
walk on the cylinder $C_L=(\Z/L\Z)\times\Z$; because $L$ is even the HV type of
Proposition~\ref{prop:hv} descends to $C_L$. For an undirected cylinder edge, its
\emph{multiplicity} is the number of traversals by the projected one-period walk.

\begin{lemma}[Visit count]\label{lem:visit}
Every vertex $v$ of $C_L$ is visited $n_v=2k_v$ times in one quotient period, with
exactly $k_v$ turns $\turnR$ and $k_v$ turns $\turnL$.
\end{lemma}

\begin{proof}
Let $z$ be a physical representative of $v$. Each phase $j$ with $p_j\in v+\Z\drift$
has a unique level $m_j$ with $p_j=z+m_j\drift$, and phase $j$ in period $-m_j$ visits
$z$; conversely every visit to $z$ over the bi-infinite orbit arises once this way.
Hence the visits to $v$ in one quotient period biject with the visits to $z$ over the
translated bi-infinite orbit. At phase $nN+j$ the black pattern is the finite translate
$B_j+n\drift$, so $z$ is white for all sufficiently early and all sufficiently
late times. Its turn sequence therefore alternates, begins with $\turnR$, ends with
$\turnL$, and contains equally many of each.
\end{proof}

At a vertex $v$ let $H_{v,-},H_{v,+}$ be the multiplicities of the west and east
horizontal edges and $V_{v,-},V_{v,+}$ those of the south and north vertical edges.
By Proposition~\ref{prop:hv} one pair carries all arrivals and the other all
departures, and each visit is one arrival plus one departure, so
\begin{equation}\label{eq:balance}
H_{v,-}+H_{v,+}=V_{v,-}+V_{v,+}=2k_v .
\end{equation}

\begin{lemma}[Edge parities]\label{lem:edgeparity}
The horizontal-edge multiplicity has a constant parity $h_y\in\F_2$ along each row
$y$. Every vertical-edge multiplicity is even; write $V_e=2G_e$ with $G_e\in\Z_{\ge0}$.
\end{lemma}

\begin{proof}
Reducing the horizontal half of \eqref{eq:balance} modulo $2$ gives
$H_{v,-}\equiv H_{v,+}$, so consecutive horizontal edges in a row share parity;
call it $h_y$. The vertical half gives $V_{v,-}\equiv V_{v,+}$, so vertical parity is
constant up each column. The projected walk has finite vertical support, so a
vertical edge far enough below or above it has multiplicity $0$; constancy then forces
every vertical multiplicity even.
\end{proof}

\begin{lemma}[Local parity]\label{lem:localparity}
At every vertex $v=(x,y)$ of $C_L$, $\;k_v\equiv h_y\pmod2$.
\end{lemma}

\begin{proof}
Suppose $v$ is vertical (the horizontal case is the $90^\circ$ rotation). By
Proposition~\ref{prop:hv} the ant arrives at $v$ along a vertical edge---from the
south moving north, or from the north moving south---and departs along a horizontal
edge. Record each visit by (arrival direction, turn). Using
$u_{q+\varepsilon}$ from \eqref{eq:step}: an arrival from the south ($\turnR$) departs
east, from the south ($\turnL$) departs west, from the north ($\turnR$) departs west,
from the north ($\turnL$) departs east. Let $a_{\turnR},a_{\turnL}$ count arrivals
from the south with the two turns and $b_{\turnR},b_{\turnL}$ those from the north.
Because the east edge is only ever traversed \emph{out} of $v$ (its other endpoint is
horizontal, hence cannot depart along it), its multiplicity equals the number of east
departures:
\[
H_{v,+}=a_{\turnR}+b_{\turnL}.
\]
The north edge is only ever traversed \emph{into} $v$, with multiplicity
$V_{v,+}=b_{\turnR}+b_{\turnL}$. The total turn count is $k_v=a_{\turnR}+b_{\turnR}$.
Hence
\[
H_{v,+}-k_v=(a_{\turnR}+b_{\turnL})-(a_{\turnR}+b_{\turnR})=b_{\turnL}-b_{\turnR}
\equiv b_{\turnL}+b_{\turnR}=V_{v,+}\pmod2 .
\]
By Lemma~\ref{lem:edgeparity}, $H_{v,+}$ has parity $h_y$ and $V_{v,+}$ is even; therefore
$h_y\equiv k_v\pmod2$.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:evenwinding}]
Halving the vertical half of \eqref{eq:balance} and using $V_e=2G_e$ gives the exact
identity $k_{x,y}=G_{x,y-1/2}+G_{x,y+1/2}$, where $G_{x,y\pm1/2}$ denote the halved
multiplicities of the vertical edges below and above $v$. Fix a column $x$ and sum
Lemma~\ref{lem:localparity} over all rows:
\begin{equation}\label{eq:telescope}
\sum_y h_y\ \equiv\ \sum_y k_{x,y}\ =\ \sum_y\bigl(G_{x,y-1/2}+G_{x,y+1/2}\bigr)\ \equiv\ 0\pmod2,
\end{equation}
because each interior half-edge $G$ appears exactly twice and the two exterior terms
vanish by finite vertical support.

Now cut $C_L$ along a vertical seam between two adjacent columns. The algebraic
winding number $w$ of the projected closed walk is the signed number of seam
crossings; there is exactly one horizontal seam edge in each row $y$, of multiplicity
$\equiv h_y\pmod2$, and signs vanish modulo $2$, so
\[
w\equiv\sum_y h_y\equiv0\pmod2
\]
by \eqref{eq:telescope}. On the other hand the lift of the closed walk runs from $p$
to $p+(L,0)$, so it winds exactly once around $C_L$: $w=1$. This contradiction proves
the theorem for drift $(L,0)$; quarter-turn rotations (including a half-turn) cover the
remaining axis-aligned drifts.
\end{proof}

\begin{corollary}[Growth is a positive multiple of four]\label{cor:growth4}
Every finite-support periodic highway has net black growth
\begin{equation}\label{eq:growth4}
\growth\in4\Z_{>0}.
\end{equation}
In particular every finite-support periodic highway builds a nonempty wake.
\end{corollary}

\begin{proof}
Zero drift confines the ant to the finitely many positions of one period, excluded by
Theorem~\ref{thm:unbounded}. Negative growth cannot recur forever since $b_t\ge0$. A
heading-resetting period has $\growth\equiv0\pmod4$ (Lemma~\ref{lem:invariant}). If
$\growth=0$ then by Corollary~\ref{cor:sep} the state reduces, after deleting a finite
stationary set never revisited, to a clean finite translator with nonzero drift,
contradicting Theorem~\ref{thm:evenwinding}. Hence $\growth>0$, and
$\growth\in4\Z_{>0}$.
\end{proof}

The standard highway has $\growth=12$, consistent with \eqref{eq:growth4}. The
telescope \eqref{eq:telescope} is the finite algebraic core machine-checked in Lean
(Appendix~\ref{app:lean}).

%=======================================================================
\section{The signed mod-four wake-residue theorem}\label{sec:residue}
%=======================================================================

We now give a second, independent proof that no zero-growth periodic highway exists,
which additionally pins the arithmetic relation between growth and drift. Recall the
strand bases $c_1,\dots,c_\growth$ of Proposition~\ref{prop:decomp}.

\begin{theorem}[Signed mod-four wake-residue identity]\label{thm:residue}
Let a finite-support repeating trace have restored heading, nonzero drift
$\drift=(a,b)$, growth $\growth$, and decomposition \eqref{eq:decomp}. Then:
\begin{enumerate}[label=\textup{(\roman*)}]
\item if $a\neq0$, then for every residue $r\in\Z/|a|\Z$,
\begin{equation}\label{eq:res-a}
\sum_{i:\;c_{i,x}\equiv r\ (\mathrm{mod}\ |a|)}\che(c_i)\ \equiv\ 2\pmod4;
\end{equation}
\item if $b\neq0$, the analogous sum over each residue of $c_{i,y}$ modulo $|b|$ is
$\equiv2\pmod4$;
\item if $a=0$, the signed sum $\sum_{i:\,c_{i,x}=x_0}\che(c_i)$ over each exact
column $x_0$ is $\equiv0\pmod4$; symmetrically for exact rows when $b=0$.
\end{enumerate}
\end{theorem}

The proof rests on a monodromy computation for the potentials of
Proposition~\ref{prop:cocycle}.

\begin{lemma}[Diagonal increment and vertical periodicity]\label{lem:diag}
Fix $a\neq0$ and any $\alpha:\Z/|a|\Z\to\Z/4\Z$, and set $w(x,y)=\che(x,y)\,\alpha(x)$,
a weight valued in $\Z/4\Z$. Then $w$ satisfies the four-corner condition
\eqref{eq:cocycle} in $A=\Z/4\Z$, so potentials $F_q$ as in
\eqref{eq:pot} exist, and
\begin{equation}\label{eq:diag}
F_N(x{+}1,y{+}1)-F_N(x,y)=-\che(x,y)\bigl(\alpha(x)+\alpha(x{+}1)\bigr),
\end{equation}
while $F_N$ is $2$-periodic in $y$.
\end{lemma}

\begin{proof}
Since $\che(x{+}1,y{+}1)=\che(x,y)$ and $\che(x{+}1,y)=\che(x,y{+}1)=-\che(x,y)$, the
four-corner sum \eqref{eq:cocycle} telescopes to
$\che(x,y)[\alpha(x{+}1)-\alpha(x{+}1)-\alpha(x)+\alpha(x)]=0$;
so $F_q$ exist by Proposition~\ref{prop:cocycle}. Compose the $\turnR$ step at $(x,y)$ heading $N$ (giving
$F_E(x{+}1,y)-F_N(x,y)=-w(x,y)$) with the $\turnL$ step at $(x{+}1,y)$ heading $E$
(giving $F_N(x{+}1,y{+}1)-F_E(x{+}1,y)=+w(x{+}1,y)$). Adding,
\[
F_N(x{+}1,y{+}1)-F_N(x,y)=-w(x,y)+w(x{+}1,y)=-\che(x,y)\alpha(x)-\che(x,y)\alpha(x{+}1),
\]
which is \eqref{eq:diag}. The relation $F_S=F_N-2w$ (from composing the two vertical
steps) gives the same increment along $(1,-1)$; combining the $(1,1)$ and $(1,-1)$
increments over a closed square shows $F_N(x,y{+}2)=F_N(x,y)$.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:residue}]
Fix $a\neq0$, $\alpha$, and $w$ as in Lemma~\ref{lem:diag}. Restored heading gives
$\growth\equiv0\pmod4$, so the period is even and $a+b$ is even; hence translation by
$\drift$ preserves $\che$, and $w(z+\drift)=w(z)$. Let $\varepsilon=\operatorname{sgn}(a)$,
and set $\sigma=\operatorname{sgn}(b)$ if $b\neq0$ and $\sigma=1$ if $b=0$. Move from
$z$ to $z+\drift$ by $|a|$ diagonal steps of vector $(\varepsilon,\sigma)$, followed by
the vertical displacement $(0,\,b-\sigma|a|)$, which is even because $a+b$ is even. The
uniform diagonal increment of $F_N$ (read off from \eqref{eq:diag} for
$\varepsilon=+1$ and from its reverse for $\varepsilon=-1$) is
\begin{equation}\label{eq:uniform}
F_N(x+\varepsilon,y+\sigma)-F_N(x,y)=-\varepsilon\,\che(x,y)\bigl(\alpha(x)+\alpha(x+\varepsilon)\bigr),
\end{equation}
and $\che$ is constant along the path because $\varepsilon+\sigma$ is even; the
vertical part contributes nothing by the $2$-periodicity of $F_N$ in $y$. As $x$ runs
once through $\Z/|a|\Z$ each $\alpha(r)$ appears twice, so the increments telescope to
$-2\varepsilon\,\che(z)\sum_{r}\alpha(r)$, giving the monodromy
\begin{equation}\label{eq:monodromy}
F_N(z+\drift)-F_N(z)\ \equiv\ 2\sum_{r\in\Z/|a|\Z}\alpha(r)\pmod4
\end{equation}
(since $-2\varepsilon\che(z)\equiv2\pmod4$ for either sign of $a$). The three other headings satisfy the same relation: $F_S=F_N-2w$
and $F_E,F_W$ differ from $F_N$ by single applications of \eqref{eq:pot}, all of which
are themselves $\drift$-periodic because $w$ is; so \eqref{eq:monodromy} holds with
$F_q$ in place of $F_N$ for every $q$, and for either sign of $a$.

Charge conservation of $\sum_{z\in B}w(z)+F_q(p)$ over one period gives
\begin{equation}\label{eq:conservation}
\sum_{z}D(z)\,w(z)+F_q(p+\drift)-F_q(p)=0,
\end{equation}
where $D$ is the signed toggle pattern of Proposition~\ref{prop:decomp}. Because $w$
is $\drift$-periodic and $(T_\drift A)(z)=A(z-\drift)$, the widget terms cancel:
$\sum_z(T_\drift A)(z)w(z)=\sum_zA(z)w(z)$. Substituting $D=S+(T_\drift-1)A$ and the
monodromy \eqref{eq:monodromy} into \eqref{eq:conservation},
\begin{equation}\label{eq:reduced}
\sum_{i=1}^{\growth}\che(c_i)\,\alpha(c_{i,x})+2\sum_{r\in\Z/|a|\Z}\alpha(r)\ \equiv\ 0\pmod4 .
\end{equation}
Taking $\alpha=\mathbf 1_{\{r\}}$ turns \eqref{eq:reduced} into
$\sum_{i:\,c_{i,x}\equiv r}\che(c_i)+2\equiv0$, i.e.\ \eqref{eq:res-a}, using
$-2\equiv2$. Rotating by $90^\circ$ gives (ii). For $a=0$, take a finitely supported
exact-column $\alpha:\Z\to\Z/4\Z$; then $b$ is even, $w$ is $\drift$-periodic, the
vertical $2$-periodicity makes the monodromy vanish, and exact-column indicators give
(iii).
\end{proof}

\begin{corollary}[Strand-density bound]\label{cor:P16}
Every finite-support repeating trace with drift $\drift=(a,b)$ has
\begin{equation}\label{eq:P16}
\growth\ \ge\ 2\max(|a|,|b|).
\end{equation}
In particular $\growth=0$ is impossible---an independent proof of
Theorem~\ref{thm:evenwinding}. If a residue fiber contains exactly two strand bases,
they share checkerboard parity; for $\drift=(\pm2,\pm2)$ and $\growth=4$ the
$2\times2$ residue-count matrix is $\left[\begin{smallmatrix}2&0\\0&2\end{smallmatrix}\right]$
or $\left[\begin{smallmatrix}0&2\\2&0\end{smallmatrix}\right]$.
\end{corollary}

\begin{proof}
A sum of $m$ signs $\pm1$ congruent to $2\pmod4$ is nonzero, with $m$ positive and
even. Hence each of the $|a|$ residue classes modulo $|a|$ contains at least two
strand bases, giving $\growth\ge2|a|$; symmetrically $\growth\ge2|b|$. If $\growth=0$
some class is empty while $\drift\neq0$, contradicting \eqref{eq:res-a}. A two-element
fiber has $\che$-sum $\pm2$, forcing equal signs. When
$\drift=(\pm2,\pm2)$ and $\growth=4$, every row and column of the $2\times2$ residue
matrix therefore has sum $2$. Equal checkerboard signs in either fixed-$x$ row force
its two bases into the same $y$-parity column, ruling out the matrix with all entries
$1$; the only possibilities are the two matrices displayed in the statement.
\end{proof}

\begin{remark}
The bound \eqref{eq:P16} is a genuine lower bound but is not claimed to be attained by
a chronological highway: the standard highway has $\growth=12$ against
$2\max(2,2)=4$. Determining which residue profiles can be realised by a single ant
tour is part of the chronology problem left open in Section~\ref{sec:gap}.
\end{remark}

Unlike the even-winding argument, Theorem~\ref{thm:residue} uses neither axis
alignment nor the clean-packet reduction; it applies to every repeating trace and
additionally fixes the drift--growth arithmetic. Independent computation supports the
theorem in the ranges where it is testable (Appendix~\ref{app:repro}): the underlying
toggle-charge identity \eqref{eq:conservation} was confirmed on all $172{,}092$
heading-reset nonzero-drift words through length $18$. The strand identity
\eqref{eq:res-a} and the bound \eqref{eq:P16} were confirmed on the standard highway,
its powers, quarter-turn rotations, and cyclic phase shifts; the direct strand-level
test is vacuous through
length $18$ because no realisable nonzero-drift trace occurs in that range.

%=======================================================================
\section{Collision chains}\label{sec:interlace}
%=======================================================================

We encode a finite cell set as edges of the bipartite row--column incidence graph:
the cell $(x,y)$ is the edge $C_xR_y$ between a column vertex $C_x$ and a row vertex
$R_y$. To a directed state $s=(p,q)$ assign the token $\eta(s)=C_{p_x}$ if $q$ is
vertical and $\eta(s)=R_{p_y}$ if $q$ is horizontal.

\begin{theorem}[Two-endpoint collision-chain parity]\label{thm:collision}
Over $\F_2$, the boundary (set of odd-degree vertices) of the row--column incidence
graph of the symmetric difference $B_s\triangle B_t$ equals $\eta(s)+\eta(t)$.
Intermediate tokens telescope across any collision checkpoints; consequently at most
one defect component of $B_s\triangle B_t$ is non-Eulerian, and a same-pose return
($\eta(s)=\eta(t)$) has only Eulerian defect components.
\end{theorem}

\begin{proof}
One step toggles the single edge $C_{p_x}R_{p_y}$, changing the degree parity of
exactly $C_{p_x}$ and $R_{p_y}$. If the pre-step heading is vertical then
$\eta(s)=C_{p_x}$ and the post-step heading is horizontal with unchanged row, so
$\eta(s')=R_{p_y}$; for a horizontal pre-step heading the roles are reversed. Hence
$\eta(s)+\eta(s')=C_{p_x}+R_{p_y}$, exactly the one-step boundary contribution.
Summing over a run telescopes to
$\eta(s)+\eta(t)$. By the handshaking lemma, every connected component has an even
number of odd-degree vertices. If $\eta(s)\ne\eta(t)$, the entire graph has exactly
those two odd vertices, so they lie in the same component; that component is the
unique non-Eulerian one. If $\eta(s)=\eta(t)$, the boundary is empty and every
component is Eulerian.
\end{proof}

This yields a four-cell lower bound on completed lifetimes. Take $s$ immediately after
an $\turnR$ visit to $z=(x,y)$ and $t$ immediately before its next ($\turnL$) visit.
Then $z$ is black at both times, so its edge $C_xR_y$ is absent from $G:=B_s\triangle
B_t$. If arrivals at $z$ are vertical, the $\turnR$ step leaves $z$ horizontally
along row $y$, so $\eta(s)=R_y$, while $\eta(t)=C_x$; if arrivals are horizontal the
roles are reversed. Thus in either case the tokens are the two ends of the missing
edge: $\eta(s)+\eta(t)=C_x+R_y$. By Theorem~\ref{thm:collision},
$\partial G=C_x+R_y$, so $C_x$ and $R_y$ are joined in $G$ by a path avoiding the edge
$C_xR_y$; in the simple bipartite incidence graph such a path has at least three edges.
Adjoining $C_xR_y$ closes a cycle of length at least four, so the completed
$\turnR$-to-$\turnL$ lifetime involves at least four distinct cells.

Circuit-partition and interlacement formalisms may provide a useful compression of
the chronological constraints, but we make no touch-graph rank claim here. The
two-endpoint boundary identity above is the only result of this section used later.

%=======================================================================
\section{Transverse rigidity and period-unbounded exclusions}\label{sec:transverse}
%=======================================================================

Every exclusion obtained so far is bounded: Theorem~\ref{thm:periodexcl} below rules
out periods up to a fixed cutoff, and the residue identity constrains arithmetic
rather than geometry. This section gives three statements that hold at \emph{every}
period. Extremal rigidity and the width-two theorem use the ordering content of
Theorem~\ref{thm:criterion}---condition (ii), that the stabilised word begins with
$\turnR$, and condition (i), that it alternates.  The computer-assisted width-four
theorem then combines extremal rigidity with exact local ant dynamics.  None uses
Section~\ref{sec:residue}.

Throughout this section the drift is \emph{diagonal}: $\drift=(a,b)$ with
$|a|=|b|=m\ge1$. A quarter-turn rotation of the plane sends $\drift=(a,b)$ to
$(-b,a)$, hence sends $ab$ to $-ab$, and commutes with the ant's rule; since
Definition~\ref{def:periodic} identifies highways up to quarter-turn rotation, we may
and do \emph{normalise so that $ab<0$}, and then set
\begin{equation}\label{eq:tfun}
t(x,y)=x+y,
\end{equation}
so that $\drift=m e$ with $e=\pm(1,-1)$. The standard highway is already in this
frame, with $\drift=(2,-2)$, $m=2$ and $e=(1,-1)$.

With this normalisation $t(\drift)=0$, so $t$ is constant on translation classes and
the range of $t$ along the trace is the same in every period; and \emph{every unit
step changes $t$ by exactly $\pm1$}, the four increments being $+1$ for east and
north and $-1$ for west and south. Diagonality is what supplies the latter: for an
axis drift the corresponding transverse functional is $x$ or $y$, which two of the
four moves leave unchanged, and the arguments below fail.

The normalisation is not cosmetic. In the rotated frame $ab>0$ one must instead take
$t=x-y$, whose increments are $+1$ for east and \emph{south} and $-1$ for west and
\emph{north}; the roles of the two axes are exchanged, and every horizontal/vertical
assertion below is exchanged with it. Only the conclusions that do not name an axis
---that every extremal turn is $\turnR$, that extremal cells are visited once, and
that $W$ is even---are stated the same way in both frames. Remark~\ref{rem:frames}
records the unnormalised form.

Put
\[
T=\max_i t(p_i),\qquad T'=\min_i t(p_i),\qquad W=T-T',
\]
and call $W$ the \emph{transverse width}. Call a lattice cell \emph{horizontal} or
\emph{vertical} according to whether the arrivals at it are east/west or north/south;
by Proposition~\ref{prop:hv} this is determined by its checkerboard parity. Since all
cells on a level line $t=c$ share the parity of $c$, each level line is entirely
horizontal or entirely vertical.

\begin{theorem}[Extremal-line rigidity]\label{thm:extremal}
Let $w^\omega$ be a finite-support periodic highway with diagonal drift, normalised
as above so that $t=x+y$. Then
\begin{enumerate}[label=\textup{(\roman*)}]
\item both extremal level lines $t=T$ and $t=T'$ are horizontal;
\item every turn on those two lines is $\turnR$; the arrival heading is east on the
line $t=T$ and west on the line $t=T'$;
\item every cell of those two lines that the periodic trace visits at all is visited
\emph{exactly once}; equivalently, every translation class meeting an extremal line is
a singleton, so each such cell is entered once, is left black, and is never returned
to;
\item $W$ is even.
\end{enumerate}
\end{theorem}

\begin{proof}
Let $I$ be a translation class meeting the line $t=T$, and consider its cells
$z_n=b_I+n\drift$; since $t(\drift)=0$ every $z_n$ lies on that same line. Take $n$
large enough that $z_n$ has been reached by every phase of $I$, so that the
chronological turn sequence at $z_n$ is the stabilised word of $I$; in particular
every visit to $z_n$ has a predecessor. The ant can reach $t=T$ only from $t=T-1$, so
each arrival at $z_n$ increases $t$; and it can never occupy $t>T$, so no departure
from $z_n$ increases $t$.

We use the explicit turn images: $\turnR$ is clockwise, so it sends north, east,
south, west to east, south, west, north; $\turnL$ is anticlockwise and reverses this.
With $t=x+y$ the increments are $+1$ for east and north and $-1$ for west and south.

Consider $t=T$. The arrival at $z_n$ increases $t$, so it is east or north, and the
departure must decrease $t$. If the line were vertical the arrival would be north,
whose clockwise image is east---which increases $t$ and is therefore forbidden---while
its anticlockwise image is west, which is allowed; every turn at $z_n$ would then be
$\turnL$, so the stabilised word of $I$ would begin with $\turnL$, contradicting
Theorem~\ref{thm:criterion}(ii). Hence the line is horizontal and the arrival is east.
Now the anticlockwise image of east is north, which increases $t$ and is forbidden,
while its clockwise image is south, which is allowed; so every turn at $z_n$ is
$\turnR$.

At $t=T'$ the arrival decreases $t$, so it is west or south, and the departure must
increase $t$. If the line were vertical the arrival would be south, whose clockwise
image is west decreases $t$ and is forbidden, while its anticlockwise image east is
allowed; every turn would be $\turnL$, the same contradiction. Hence that line is
horizontal too, the arrival is west, its clockwise image north is allowed and its
anticlockwise image south is forbidden, so again every turn is $\turnR$. This proves
(i) and (ii).

In both cases the stabilised word of $I$ is a block of $\turnR$'s, and by
Theorem~\ref{thm:criterion}(i) it alternates, so that block has length one: $I$ is a
singleton. A singleton class is visited once at each of its cells and is odd, so every
cell of $I$ that the trace reaches is entered exactly once and left black. As $I$ was
an arbitrary class meeting an extremal line, this proves (iii). Both extremal lines
being horizontal, they carry the same checkerboard parity, so $T\equiv T'\pmod2$,
which is (iv).
\end{proof}

Theorem~\ref{thm:extremal} says that a periodic highway is bounded transversally by
two guard rails of permanent wake, touched once and never returned to. The mechanism
deserves a remark, since it is what the incidence arguments of the preceding sections
cannot reach: the vertical case is eliminated purely by the \emph{first-turn}
condition, and the single visit purely by \emph{alternation}. Both are assertions
about the order in which a cell is visited, and the countermodels recorded in
Section~\ref{sec:gap} satisfy every residue equation while violating the conclusion,
so neither is a consequence of the arithmetic developed there.

\begin{remark}[the unnormalised statement]\label{rem:frames}
Without the normalisation $ab<0$ the conclusion must be read in the frame determined
by the drift, and it is easy to state it wrongly. In the rotated frame $ab>0$ one
takes $t=x-y$; the headings increasing $t$ are then east and \emph{south}, the
argument of Theorem~\ref{thm:extremal} eliminates the \emph{horizontal} case, and both
extremal lines are \emph{vertical}, with arrival south on $t=T$ and north on $t=T'$.
What is common to the two frames, and is what the sequel uses, is that the two
extremal lines have the \emph{same} type, that every extremal turn is $\turnR$, that
extremal cells are entered at most once, and that $W$ is even. Concretely, the
standard highway read from a north-facing start has $\drift=(2,-2)$ and horizontal
extremal lines with arrivals east and west; the same highway read from an east-facing
start---the same object, rotated---has $\drift=(-2,-2)$ and \emph{vertical} extremal
lines with arrivals south and north. In both cases every extremal turn is $\turnR$.
\end{remark}

The rigidity is strong enough to collapse a narrow strip to a one-dimensional system,
which can then be refuted at all periods simultaneously.

\begin{theorem}[No highway of transverse width two]\label{thm:width2}
No finite-support periodic highway with diagonal drift has transverse width $W=2$.
Combined with Theorem~\ref{thm:extremal}(iv), every such highway has even width
$W\ge4$.
\end{theorem}

\begin{proof}
Suppose $W=2$ and label the three occupied level lines $0,1,2$, meaning $t=T,\,T-1,\,
T-2$. By Theorem~\ref{thm:extremal} lines $0$ and $2$ are horizontal, each of their
cells is entered at most once and turns $\turnR$; line $1$ is vertical. Because each
step changes $t$ by $\pm1$, the ant alternates between line $1$ and lines $\{0,2\}$,
so consecutive visits to line $1$ are separated by exactly one excursion to line $0$
or to line $2$.

Theorem~\ref{thm:extremal} makes those excursions rigid. Orient the longitudinal axis
so that $e=(1,-1)$; recall east and north increase $t$. From a cell $z$ on line $1$
the upward excursion
enters line $0$ heading east and leaves it heading south, so it carries the ant from
$z$ to $z+(1,0)+(0,-1)=z+e$; the downward excursion enters line $2$ heading west and
leaves heading north, carrying $z$ to $z-e$. Index the cells of line $1$ by
$\ell\in\Z$ along $e$. Every excursion changes $\ell$ by exactly $\pm1$.

Now bound how often the ant can cross between adjacent sites. An upward excursion
from site $\ell$ passes through the line-$0$ cell determined by $\ell$; call it
$U_\ell$, and let $D_\ell$ be the line-$2$ cell used by a downward excursion from
$\ell$. Distinct sites give distinct cells, and by Theorem~\ref{thm:extremal}(iii)
each of $U_\ell$ and $D_\ell$ is entered at most once by the whole periodic trace.
Since the gap between sites $\ell$ and $\ell+1$ is crossed upward only by an upward
excursion from $\ell$, and downward only by a downward excursion from $\ell+1$,
\begin{equation}\label{eq:edgebudget}
\text{each gap is crossed at most once upward and at most once downward.}
\end{equation}

Choose a cyclic phase at a visit to line $1$, and let $s_0$ be that initial site.
Because the drift is $+m$ along
$\ell$, the ant's site tends to $+\infty$; so for every $\ell\ge s_0$ the walk starts
weakly left of the gap $(\ell,\ell+1)$ and is eventually permanently to its right,
whence its net number of upward minus downward crossings of that gap is exactly $1$.
By \eqref{eq:edgebudget} the only way to achieve a net of $1$ with each count at most
$1$ is one upward crossing and \emph{no} downward crossing. Hence for every
$\ell>s_0$ the ant enters site $\ell$ exactly once, necessarily from $\ell-1$, and
leaves it towards $\ell+1$, never returning.

It remains to read off the turn at such a site. Arriving from $\ell-1$ means
completing an upward excursion, which leaves line $0$ heading south; so the ant
arrives at site $\ell$ heading south. Departing towards $\ell+1$ means beginning an
upward excursion, which leaves site $\ell$ heading east. The anticlockwise image of
south is east and the clockwise image is west, so the unique turn at site $\ell$ is
$\turnL$.

Finally, let $I$ be the translation class of such a site with $\ell>s_0$. Its cells
are the sites $\ell+nm$, all but finitely many of which exceed $s_0$ and are therefore
entered exactly once. So $I$ has a single phase, its stabilised word is the single
letter $\turnL$, and Theorem~\ref{thm:criterion}(ii) is violated. Since there are
infinitely many such sites, the contradiction is unavoidable.
\end{proof}

The next case admits a different reduction.  We give the finite computation as a
separate lemma so that the computer-assisted boundary of the theorem is explicit.
Write a finite sequence such as $(2,0,2)$ compactly as $202$.

\begin{lemma}[blank-column classification]\label{lem:blankcolumn}
Consider an initially white column consisting of the five cells on transverse
lines $0,1,2,3,4$.  At each of its left and right boundaries record, in chronological
order, the transverse line of the destination cell immediately after every horizontal
crossing.  Suppose each sequence starts and ends with a rightward crossing, its
directions alternate, every rightward label belongs to $\{2,4\}$, every leftward
label belongs to $\{0,2\}$, and on each boundary the rightward label $4$ and
leftward label $0$ occur at most once.  The compatible ordered pairs of left and
right sequences are exactly the following twelve:
\[
\begin{array}{c|l}
\text{left sequence}&\text{possible right sequences}\\ \hline
202&4\\
4220222&2,\ 202,\ 20222,\ 2022224\\
2042222&2,\ 202,\ 20222,\ 2022224\\
202242222&422,\ 42202,\ 4220222.
\end{array}
\]
The directed graph formed by these pairs is acyclic; its longest directed path has
three edges.
\end{lemma}

\begin{proof}[Exact finite verification]
A column state consists of its five colour bits, the transverse entry line and
heading, and four Boolean flags recording whether the exceptional top-rightward or
bottom-leftward event has already occurred on either boundary.  Starting from the
all-white mask and a first rightward entry on line $2$ or $4$, literal application of
the ant rule reaches $22$ such entry states, has no reachable state cycle and stops
after depth at most eight.  Exhausting the permitted return entry ($2$ or the one
available extreme) gives precisely the twelve pairs displayed above.

This enumeration is implemented twice.  The primary implementation uses a five-bit
integer mask and numeric headings; an independently written verifier uses a set of
black rows and compass-letter headings and imports no code from the first.  Both
return the same twelve pairs.  As a convention check, the second implementation
extracts two far-ahead crossing sequences from the documented standard-highway seed
and exactly replays the intervening initially white column.  Finally, assigning rank
$3$ to $202242222$, rank $2$ to $4220222$ and $2042222$, rank $1$ to $202$, and
rank $0$ to every other displayed vertex makes every edge strictly decrease rank.
The rank calculation and the impossibility of a four-edge path are also checked in
Lean~$4$.  The accompanying certificate records the edge table, source hashes and
the boundary between the enumerated and formalised parts.
\end{proof}

\begin{theorem}[No highway of transverse width four]\label{thm:width4}
No finite-support periodic highway with diagonal drift has transverse width $W=4$.
Consequently every such highway has even transverse width $W\ge6$.
\end{theorem}

\begin{proof}
Suppose $W=4$.  Rotate by $180^\circ$ if necessary so that the normalised drift is
$(m,-m)$ with $m>0$, shift the transverse coordinate so that
$t=x+y\in\{0,1,2,3,4\}$, and use $x$ as longitudinal coordinate.  A north or south
step remains in one column $x=k$, whereas an east or west step crosses a boundary
$x=k+\tfrac12$.

At the beginning of the periodic tail, only finitely many cells have been visited,
and the initial black support is finite.  Hence all sufficiently large longitudinal
columns are untouched and all five of their cells are white.  For such a far-ahead
boundary let $\Gamma_k$ be its complete future crossing sequence, recording the
destination transverse line.  The longitudinal position in successive periods is
$nm$ plus one of finitely many phase offsets.  Thus the ant begins to the left of the
boundary, crosses it only finitely many times, and is eventually permanently to its
right.  Horizontal crossing directions therefore alternate right, left, \ldots,
right, so $\Gamma_k$ is nonempty and has odd length.

On a rightward crossing the destination label is $2$ or $4$; on a leftward crossing
it is $0$ or $2$.  Direction is therefore implicit in the parity of the event index.
Moreover, two rightward labels $4$ at the same cut would enter the same top-extremal
cell twice, and two leftward labels $0$ would enter the same bottom-extremal cell
twice.  Both are forbidden by Theorem~\ref{thm:extremal}(iii).

The untouched column $x=k$ has left sequence $\Gamma_{k-1}$ and right sequence
$\Gamma_k$, so Lemma~\ref{lem:blankcolumn} makes
$\Gamma_{k-1}\to\Gamma_k$ an edge of its directed graph.  Infinitely many
consecutive untouched columns would give a directed path of arbitrary length, while
the lemma permits no path of four edges.  This contradiction excludes $W=4$.
Together with Theorem~\ref{thm:extremal}(iv) and Theorem~\ref{thm:width2}, it gives
$W\ge6$.
\end{proof}

\begin{remark}[why the same proof stops at width six]\label{rem:width4}
The standard highway has $W=10$.  The blank-column recursion is acyclic at width
four, but an extra interior horizontal line creates a reachable two-state local
oscillator at widths six and eight.  In one state the black rows are $\{1,2\}$ and
the ant enters row $2$ eastward; two turns change the mask to $\{1,3\}$ and exit
eastward.  A westward return on row $2$ restores the first state and exits westward.
The two reused cells alternate their turns correctly, and neither extreme is reused.
This is not a highway; it proves only that the acyclic one-column argument does not
extend unchanged.  A wider-width proof must couple columns or exploit the order of
whole translation classes.  In particular, a finite alphabet on an unbounded strip
does not by itself make the highway question finite-state.
\end{remark}

The hypotheses of this section are exactly Proposition~\ref{prop:hv} and
Theorem~\ref{thm:criterion}. In particular Theorems~\ref{thm:extremal},
\ref{thm:width2}, and~\ref{thm:width4} are independent of Section~\ref{sec:residue} and of the
computations of Section~\ref{sec:computation}.

%=======================================================================
\section{Small-period exclusion}\label{sec:computation}
%=======================================================================

All searches are exact: integer coordinates, exact finite sets, and no probabilistic
equality tests. Three search variants are used. The \emph{residue-free variant} uses no
consequence of Theorem~\ref{thm:residue} and applies the exact criterion of
Theorem~\ref{thm:criterion} to every positive-growth nonzero-drift leaf; its zero-hit
outcome is what certifies Theorem~\ref{thm:periodexcl} below, and that certification
is unconditional on the residue theorem. The \emph{strand-pruned} and
\emph{residue-pruned} variants additionally prune with Corollary~\ref{cor:P16} and
with the full identity of Theorem~\ref{thm:residue} respectively; they are faster, and
their agreement with the residue-free variant is reported as a cross-check. All three
accept the standard word. We stress that the three variants share one enumeration
framework, one normalisation, and one representation, so their mutual agreement is a
consistency check and not implementation independence; a separate verifier, written
from the statement of Theorem~\ref{thm:criterion} with a different class construction,
is described in Appendix~\ref{app:repro}. The methodology, per-shard records, and
cross-audit are described there as well.

\begin{theorem}[No periodic highway of period at most $48$]\label{thm:periodexcl}
No finite-support periodic highway of nonzero drift has period at most $48$.
Consequently any nonstandard finite-support periodic highway has even period at least
$50$ (the standard highway has period $104$; periods $50,52,\dots,102$ are not
addressed here).
\end{theorem}

\begin{proof}[Verification]
Odd periods restore no heading (an odd number of quarter turns is never
$0\bmod4$), and Corollary~\ref{cor:growth4} excludes zero growth at every period, so
it suffices to search even periods for positive-growth traces. The evidence is:
\begin{enumerate}[label=\textup{(\alph*)}]
\item a complete first-$\turnR$, cyclic-phase-normalised enumeration through
period $32$ ($46{,}185{,}421$ nodes at depths $6$--$32$ over $32$ prefix shards,
plus a $59$-node unprefixed baseline through depth $5$; zero highways). Every
positive-growth heading-resetting word contains $\turnR$, so a cyclic shift places
it in this enumeration; shifting only changes the chosen phase on the same periodic
orbit and preserves finite-support realisability. No growth or endpoint pruning is
applied, so the search is residue-free by construction;
independently, a brute-force Python verifier (Appendix~\ref{app:repro}) reproduces the
exclusion over all $\sum_{L=1}^{18}2^L=524{,}286$ nonempty words of length at most $18$;
\item rank-audited positive-growth searches at periods $34,\dots,48$ with the
residue-free variant, which uses no consequence of Theorem~\ref{thm:residue} and applies
the exact criterion of Theorem~\ref{thm:criterion} to \emph{every} positive-growth
nonzero-drift leaf. At the largest, period $48$, it visited $41{,}710{,}394{,}384$
nodes and reached $7{,}446{,}719{,}550$ such leaves over all $2^{16}$ prefix ranks with
no node cap, evaluated the exact criterion at every one of them, and returned zero
certificates. Every prefix rank is covered exactly once, every shard reports
completion, and the aggregate counters equal the sums of the per-rank counters
(Table~\ref{tab:excl}).
\end{enumerate}
The zero-hit outcome of the residue-free variant is the certificate of nonexistence, and
it does not assume Theorem~\ref{thm:residue}. The strand-pruned and residue-pruned
engines of Appendix~\ref{app:repro}, which do assume it, return zero certificates on
the strictly smaller trees they explore; that agreement is a cross-check only. It is,
however, evidence \emph{for} Corollary~\ref{cor:P16}: the residue-free variant reaches
the leaves P16 discards---at period $48$, $6{,}994{,}756{,}680$ of them---and finds no
realisable word among them, so within this range the pruning discarded nothing.
Table~\ref{tab:excl} summarises. Together these exclude every finite-support periodic
highway of nonzero drift and period at most $48$.
\end{proof}

\begin{table}[ht]
\centering
\begin{tabular}{@{}lrrrr@{}}
\toprule
Period & Nodes & Leaves & Criterion evaluations & Highways \\
\midrule
$\le 32$ & $46{,}185{,}480$ & --- & --- & $0$\\
$34$ & $40{,}418{,}033$      & $6{,}989{,}418$      & $6{,}989{,}418$      & $0$\\
$36$ & $108{,}995{,}287$     & $18{,}928{,}333$     & $18{,}928{,}333$     & $0$\\
$38$ & $293{,}763{,}689$     & $51{,}386{,}767$     & $51{,}386{,}767$     & $0$\\
$40$ & $791{,}801{,}484$     & $139{,}000{,}900$    & $139{,}000{,}900$    & $0$\\
$42$ & $2{,}133{,}302{,}151$ & $376{,}781{,}423$    & $376{,}781{,}423$    & $0$\\
$44$ & $5{,}748{,}138{,}964$ & $1{,}018{,}293{,}035$& $1{,}018{,}293{,}035$& $0$\\
$46$ & $15{,}483{,}269{,}352$& $2{,}757{,}152{,}898$& $2{,}757{,}152{,}898$& $0$\\
$48$ & $41{,}710{,}394{,}384$& $7{,}446{,}719{,}550$& $7{,}446{,}719{,}550$& $0$\\
\bottomrule
\end{tabular}
\caption{Exhaustive exclusion of finite-support periodic highways of nonzero drift.
All rows from period $34$ up are the output of the \emph{residue-free} variant, which
assumes no consequence of Theorem~\ref{thm:residue}; the ``criterion evaluations''
column equals the leaf column, i.e.\ the exact criterion of
Theorem~\ref{thm:criterion} is applied to every positive-growth nonzero-drift leaf,
with no leaf discarded unexamined. Each period-$34$--$48$ entry is complete over its
parameter range with no node cap, and every prefix rank is covered exactly once. The
period-$\le32$ row enumerates all first-$\turnR$ cyclic-phase representatives, plus
the unprefixed depths below six. Every positive-growth heading-resetting word is
represented, and no growth or endpoint pruning is applied, so this row is
residue-free by construction. Its prefix shards were configured with a
$10^8$-node safety cap, but every shard completed below that cap (the largest used
$2{,}623{,}020$ nodes). Node counts are not comparable across different prefix
lengths (Appendix~\ref{app:repro}).}
\label{tab:excl}
\end{table}

A finite-period exclusion does not by itself establish eventual convergence for
arbitrary finite initial configurations: Theorem~\ref{thm:periodexcl} bounds the
minimal period of any nonstandard highway from below, but Conjecture~\ref{conj:main}
quantifies over all finite configurations and all periods, and no finite computation
of the present kind exhausts that range.

%=======================================================================
\section{The remaining gap}\label{sec:gap}
%=======================================================================

The results above constrain the arithmetic and topology of any finite-support
periodic highway, but do not force convergence. Two distinct obstructions delimit the
gap.

\begin{problem}[Bounded retreat / bounded active core]\label{prob:retreat}
Show that for every finite-support initial configuration there are a finite window
$W\subset\Z^2$, a time $t_0$, and translations $v_t$ such that, for every
$t\ge t_0$, the ant and every cell or Tait-boundary component capable of influencing
the future orbit lie in $v_t+W$; all components outside that translated window are
permanently inert and never reabsorbed. This is a bounded-active-core formulation,
stronger than merely bounding the ant's distance behind a record frontier.
\end{problem}

Under this formulation the active window has finitely many states modulo
translation, determinism forces eventual periodicity, and Theorem~\ref{thm:criterion}
decides whether the resulting cycle is finite-support realisable and constructs a
seed when it is. We do not claim these are the only possible routes to a
solution, but no proof along them is known, and the following show why several easy
attempts fail: unboundedness (Theorem~\ref{thm:unbounded}) permits an orbit that keeps
enlarging a two-dimensional region; no nonconstant affine combination of
$(E,V,k,\beta)$ is monotone (Proposition~\ref{rem:nomonotone});
and a local pumping argument is invalid because equal local cores may connect through
arbitrarily distant explored territory, turning a later merge into a split
(Theorem~\ref{thm:conjugacy}).

\begin{problem}[Periodic classification]\label{prob:periodic}
Show that every positive-growth periodic highway admitted by finite support agrees
in path and turn trace with the standard period-$104$ highway up to lattice
translation, quarter-turn rotation, and cyclic phase shift, as in
Definition~\ref{def:periodic}.
\end{problem}

At fixed growth $\growth=4$ the residue skeletons of Theorem~\ref{thm:residue} leave
unbounded free parameters, and explicit heading-reset countermodels satisfy every
incidence and residue condition yet
fail the cross-period criterion. A resolution therefore needs chronological input
valid for unbounded period---genuinely more than the incidence data used here.
Section~\ref{sec:transverse} supplies the first such input in this paper:
Theorems~\ref{thm:extremal} and~\ref{thm:width2} are proved from the ordering content
of Theorem~\ref{thm:criterion}, and the computer-assisted
Theorem~\ref{thm:width4} combines that rigidity with exact column dynamics.  All
three hold at every period and are not consequences of the residue equations.  They
pin the boundary of the strip and exclude widths two and four, but
Problem~\ref{prob:periodic} still requires control of all wider strips, non-diagonal
periodic traces, and uniqueness within any surviving geometry.  Remark~\ref{rem:width4}
also shows why ``fixed width is finite-state'' is not presently a theorem: at width
six the local column dynamics already has recurrent interior memory.  Widths six and
eight are concrete next targets, not claims of a completed decision procedure.
A \emph{disproof} of Conjecture~\ref{conj:main} would
instead exhibit an infinite certificate: a nonstandard periodic word passing
Theorem~\ref{thm:criterion}, or a finite self-enlarging Tait gadget whose state recurs
with a parameter increased. No such certificate has been found.

%=======================================================================
\section{Concluding remarks}\label{sec:conclusion}
%=======================================================================

We have developed a structural and arithmetic theory of the classical
finite-support periodic highway: a decidable realisability criterion, an exact
medial-graph conjugacy with surgery calculus, two independent proofs that no
zero-growth clean translator exists (so every highway grows a wake of size
$\in4\Z_{>0}$), the strand-density bound $\growth\ge2\|\drift\|_\infty$, a
two-endpoint collision-chain parity theorem, a transverse rigidity theorem with the
resulting all-period exclusions of diagonal transverse widths two and four, and an
exhaustive exclusion of all periodic highways of period at most $48$. Selected
algebraic kernels used in the even-winding and residue arguments are machine-checked
in Lean~$4$.

Three of these results are of a different character from the rest. The arithmetic
constraints are incidence statements: they count strands, edges and charges, and
Section~\ref{sec:gap} records countermodels satisfying all of them. Theorems
\ref{thm:extremal} and~\ref{thm:width2} instead use the order in which a cell is
visited, while Theorem~\ref{thm:width4} turns that rigidity into an acyclic exact
crossing-sequence graph.  They apply at every period rather than up to a search
bound.  The width-six interior oscillator shows, however, that the one-column
acyclicity mechanism ends sharply at width four.  Widths six and eight are concrete
next targets, but their unbounded longitudinal memory has not been reduced to a
finite decision problem.

One identified route to a full solution is a bounded-retreat or bounded-active-core
lemma followed by a chronology theorem for unbounded period.  The arithmetic
constraints proved here do not supply either step, and other routes are not excluded.

\subsection*{Acknowledgements}
The author thanks the maintainers of the Lean~4 community for the toolchain used in
Appendix~\ref{app:lean}. AI systems were used for exploratory computation, drafting,
and adversarial review.  Their output was treated as untrusted: promoted claims were
rechecked against the displayed arguments, independent implementations, and released
certificates.  The author is responsible for all remaining errors.

\appendix
%=======================================================================
\section{Worked example: the standard highway}\label{app:example}
%=======================================================================

The blank orbit first enters its repeating trace at step $9977$. Normalised so that
the ant faces north at the origin and the trace begins with $\turnR$ at a global
minimum of the black-count walk, the period-$104$ word (with $\turnR=$R, $\turnL=$L)
is
\begin{center}\ttfamily\small
\begin{tabular}{@{}l@{}}
RRRRLLRLLRRRRLLRRRRLLRLRRR\\
RLRLLLLRRRRLRRLRRRRLLLLRLR\\
RRRLRRRRLLLLRLRRRRLRLLRRLL\\
LLRRLLRRRRLLRRLRLLRLLRLRLL
\end{tabular}
\end{center}
(read as one word of $104$ symbols, four rows of $26$). It has
$\#\turnR-\#\turnL=12$ and normalised drift $(2,-2)$, in agreement with
Corollary~\ref{cor:growth4} ($\growth=12\in4\Z_{>0}$) and Corollary~\ref{cor:P16}
($12\ge 2\max(2,2)=4$). Applying the seed construction \eqref{eq:seed} of
Theorem~\ref{thm:criterion} to this exact word (in this exact phase and orientation)
produces the $11$-cell black seed
\[
\begin{aligned}
&(-2,-2),\,(-2,-1),\,(-1,0),\,(0,1),\,(1,-2),\,(1,1),\\
&(2,-2),\,(2,0),\,(3,-1),\,(3,0),\,(4,0),
\end{aligned}
\]
which, with a north-facing ant at the origin, produces the exact $104$-step word
above immediately, restoring the north heading and translating by $(2,-2)$; the
leading corridor is white and the gateway pattern recurs translated by the drift
each period. (Both the word and this seed were verified by direct simulation. A
different but equally standard normalisation---the phase obtained by cyclically
shifting this word by $100$ symbols, whose trace begins with $\turnL$ and has drift
$(-2,2)$---is generated by a $13$-cell seed; the two must not be interchanged, as
they are different phases of the same highway.)

%=======================================================================
\section{Lean 4 formalization}\label{app:lean}
%=======================================================================

A dependency-light Lean~$4$ project (pinned to Lean $4.32.0$, standard library only)
machine-checks the exact transition kernel and selected algebraic kernels used above.
Every audited theorem depends only on the standard axioms
\texttt{propext}, \texttt{Classical.choice}, and \texttt{Quot.sound}; there are no
\texttt{sorry}s, no custom axioms, and no \texttt{native\_decide}. The checked
statements include the finite-support step kernel and its toggle lemmas, the
heading/black-count reset of Lemma~\ref{lem:invariant}, the four-corner admissibility
and mod-four fiber arithmetic of Theorem~\ref{thm:residue} (including
``one residue identity per class $\Rightarrow\growth\ge2n$''), the checker-cycle
monodromy $2\sum\alpha$ of Lemma~\ref{lem:diag}, translated-widget cancellation, and
the collision-chain endpoint parity of Theorem~\ref{thm:collision}.  It also checks
the finite rank certificate for Lemma~\ref{lem:blankcolumn}: every one of the twelve
listed edges lowers rank, so no four-edge chain exists.  Completeness of the local
Python enumeration is explicitly outside the present Lean boundary.

The finite parity telescope \eqref{eq:telescope}---the combinatorial heart of the
even-winding theorem---is checked in the following self-contained form
(\texttt{evenWindingParityCore}); \texttt{xorAll} is the $\F_2$ sum of a list,
\texttt{incidences} takes successive XORs, and the two \texttt{false} endpoints are
the vanishing exterior half-flows:

\begin{lstlisting}[style=lean]
def bxor : Bool -> Bool -> Bool
  | false, b     => b
  | true,  false => true
  | true,  true  => false

def xorAll : List Bool -> Bool
  | []      => false
  | b :: bs => bxor b (xorAll bs)

def incidences : List Bool -> List Bool
  | []             => []
  | [_]            => []
  | a :: b :: rest => bxor a b :: incidences (b :: rest)

-- XOR of all adjacent incidences of a list bracketed by two `false`
-- endpoints is `false`: the seam-crossing count is even.
theorem evenWindingParityCore (interior : List Bool) :
    xorAll (incidences (false :: interior ++ [false])) = false := by
  rw [xorAll_incidences, lastFrom_append_singleton]; rfl
\end{lstlisting}

The formalization isolates, as an explicit interface, the endpoint normal form and
lifted potential geometry of a genuine periodic ant trace. In particular,
\texttt{StabilizedP3Word.toOrbitEndpointData} constructs all algebraic endpoint data
(first positive strand, balanced sum, and binary scan prefix) from an explicitly
supplied stabilised P3 word. What remains outside Lean is the extraction of that word
and of the local geometric charge certificates from an arbitrary periodic ant orbit.
Thus Theorem~\ref{thm:residue} is certified conditional on a structured trace
certificate rather than end to end; neither it nor Conjecture~\ref{conj:main} is
asserted as an unconditional Lean theorem. The project and build instructions
accompany the paper \cite{Repo}.

%=======================================================================
\section{Computational methodology and reproducibility}\label{app:repro}
%=======================================================================

The periodic search space is the tree of legal prefixes. The following cycle-lemma
normalisation justifies the prefix-growth pruning used by the search. Write the word
weights as $u_0,\ldots,u_{N-1}\in\{+1,-1\}$, with $\turnR$ assigned $+1$ and
$\turnL$ assigned $-1$. Define $H_0=0$ and
$H_j=\sum_{i=0}^{j-1}u_i$ for $1\le j\le N$, and suppose
$H_N=\growth>0$. Choose $j\in\{0,\ldots,N-1\}$ at which $H_j$ is minimal. For a
length-$r$ prefix of the rotation beginning at $j$, where $1\le r\le N$, the weight
is $H_{j+r}-H_j\ge0$ if $j+r\le N$, and is
$\growth+H_{j+r-N}-H_j\ge\growth>0$ if $j+r>N$. Thus every prefix of this rotation
has nonnegative weight, and its first symbol is necessarily $\turnR$. Restarting a
realisable orbit at that phase
preserves finite support, and a quarter-turn rotation restores the fixed initial
heading. Hence rejecting a prefix with negative $\turnR-\turnL$ balance loses no
positive-growth periodic trace. By Corollary~\ref{cor:growth4} only positive-growth
heading-reset words need be considered. To distribute a
period-$N$ search, the half-open interval of length-$k$ prefix ranks $[0,2^{k-1})$ is
partitioned into disjoint worker intervals; each worker performs a depth-first
enumeration of its subtree, applying the exact criterion of
Theorem~\ref{thm:criterion} at every leaf. A complete exclusion requires that every
rank be covered exactly once, that every shard report completion without reaching any
configured node cap, and
that the aggregate node, leaf, and prune counters equal the sums of the per-rank
counters.

\subsection*{What certifies the exclusion, and what merely cross-checks it}

The residue-free variant is run for every even period $34$--$48$ certification;
the two conditional pruned variants provide selected cross-checks at periods
$44$, $46$, and $48$. It matters exactly which pruning rule each one assumes, so we
spell this out.

The \emph{residue-free variant} (\texttt{PositiveGrowthSearchIndep.java}) uses only
rules that are free of Theorem~\ref{thm:residue}: the cyclic normalisation above,
same-cell alternation (Proposition~\ref{prop:alt}), the growth congruence of
Corollary~\ref{cor:growth4} (a consequence of the even-winding theorem of
Section~\ref{sec:evenwinding}, not of the residue theorem), and $L_1$ endpoint
reachability to a nonzero drift. It applies the exact
criterion of
Theorem~\ref{thm:criterion} to \emph{every} positive-growth nonzero-drift leaf. Its
zero-hit outcome is the certificate of Theorem~\ref{thm:periodexcl}, and that
certificate is unconditional on Theorem~\ref{thm:residue}. It is, however,
conditional on Corollary~\ref{cor:growth4}; we make no claim that the search is
unconditional on the even-winding theorem, which is what reduces the problem to
positive growth in the first place. Every released certifying shard has
\texttt{deficit\_depths=[]}; the optional odd-ending deficit routine is inactive, and
all of its recorded counters are zero.

The \emph{strand-pruned engine} (\texttt{PositiveGrowthSearch.java}) additionally
contracts endpoint reachability to the P16 square of Corollary~\ref{cor:P16} and
rejects leaves violating $\growth\ge2\max(|a|,|b|)$. The \emph{residue-pruned engine}
(\texttt{PositiveGrowthResidueSearch.java}) adds the full identity of
Theorem~\ref{thm:residue}. Both are therefore \emph{conditional} on the residue
theorem, and neither can certify Theorem~\ref{thm:periodexcl} on its own. Because
Corollary~\ref{cor:P16} prunes at the node level, the strand-pruned and
residue-pruned engines do not merely discard leaves: they never reach them, which is
why their reported leaf counts are smaller than the residue-free variant's rather than
equal to them, and why their agreement on node counts is a statement about the
P16-contracted tree only.

Period $46$ illustrates the scale of the conditional pruning. The residue-free
variant reaches $2{,}757{,}152{,}898$
leaves and evaluates the criterion at each; the strand-pruned engine reaches
$98{,}568{,}824$, i.e.\ $3.6\%$ of them, and the residue-pruned engine evaluates the
criterion on $6{,}132{,}192$, i.e.\ $0.22\%$. The $2{,}658{,}584{,}074$ leaves the
strand-pruned engine never reaches are exactly the ones a residue-free certificate
must inspect. Every period-exclusion count reported here is output of the residue-free
variant.

Finally, the three variants share one enumeration framework, one normalisation, one
representation, and one exact-criterion routine, so a shared bug would affect all
three identically; their agreement is a consistency check, not implementation
independence. To test the criterion itself we use a separate verifier
(\texttt{verify\_criterion\_indep.py}) written from the statement of
Theorem~\ref{thm:criterion} rather than from the Java source: it builds translation
classes by explicit pairwise membership tests and union-find instead of the Java's
arithmetic reduction and global sort, and it evaluates the \emph{literal} criterion,
checking every $S_{I,n}$ for $\min_I a_i\le n\le\max_I a_i$ rather than the single
stabilised word that Remark~\ref{rem:onesort} licenses---so it also exercises
Remark~\ref{rem:onesort}, which the Java engines assume. On the $172{,}092$
heading-reset nonzero-drift words of length at most $18$ enumerated below, the literal
criterion and the one-sort shortcut agree on every word, with zero disagreements; both
accept the standard period-$104$ word, and the Java binary agrees with the Python
verifier on every word sampled directly from it. For period at most $18$ this is a
genuine end-to-end cross-language replication: the Python verifier brute-forces all
$\sum_{L=1}^{18}2^L=524{,}286$ nonempty words with no pruning whatsoever and
independently reproduces the exclusion.
Beyond that length brute force is infeasible and the check is confined to the
criterion, which is the component a shared bug would most plausibly corrupt.

The analytic transverse results, Theorems~\ref{thm:extremal}
and~\ref{thm:width2}, are proved rather than computed, while
Theorem~\ref{thm:width4} uses the explicitly delimited finite classification in
Lemma~\ref{lem:blankcolumn}.  The analytic inputs are also checked:
\texttt{verify\_extremal.py} confirms
Theorem~\ref{thm:extremal} on the standard word (transverse range $[-5,5]$, width
$10$; each extremal line carrying two once-visited cells, all arrivals horizontal,
all turns $\turnR$), on its square, and on its $\turnR/\turnL$ reflection, where the
extremal turns are all $\turnL$ and the word is correctly rejected---the mirror image
of the vertical case excluded in the proof. More usefully it checks the
contrapositive exhaustively: of the $60{,}854$ diagonal-drift heading-resetting words
of length at most $18$, exactly $18{,}946$ have a repeated cell on an extremal line,
and \emph{none} of those is realisable, as Theorem~\ref{thm:extremal} requires.
\texttt{verify\_width.py} checks the geometric inputs of Theorem~\ref{thm:width2}---the
two forced excursions and the arrival/departure headings from which the final turn is
read---and then runs an adversarial search in which the colour of every
not-yet-visited site is chosen by an opponent rather than assumed white. With at most
$k$ black seeds no opponent survives more than $k+4$ steps before some gap must be
crossed twice in one direction, which is the budget Theorem~\ref{thm:extremal}(iii)
forbids. That the opponent is given this freedom matters: an earlier version of both
the proof and the script assumed that a site's first visit turns $\turnR$, which is
\emph{false}---condition (ii) of Theorem~\ref{thm:criterion} constrains only
stabilised words, and \eqref{eq:seed} blackens cells below their class's maximal
level precisely so that some first turns are $\turnL$. The argument given above uses
no such assumption.

For Lemma~\ref{lem:blankcolumn}, \texttt{width4\_crossing\_graph.py} supplies both
the unbounded 22-state recursion and a pairwise cross-check over all $910$ admissible
signatures through length $25$.  The independently written
\texttt{verify\_width4\_crossing\_graph\_indep.py} shares neither its state
representation nor its turn implementation, reproduces the same twelve edges, and
performs the standard-highway column replay described in the lemma.  The machine-
readable certificate \texttt{width4\_crossing\_graph\_certificate.json} contains the
edge list, statistics, source hashes and formalisation boundary.  A separate
set of combined exact searches, using the proved bound $P\le7m$, covers drift
magnitudes $1\le m\le9$ with no hit.  The fixed-drift macro engine supplies all
periods for $m\le6$ and the upper ranges for $m=7,8,9$; the general and unrestricted
width-four searches supply their lower ranges.  The $m=10$ record is explicitly
partial through $P=66$.  These searches are independent bounded cross-checks, not inputs to
Theorem~\ref{thm:width4}.

One further assumption of the engines is worth isolating, since it is what makes the
search exhaustive. The engines never test the heading directly; they test
$\growth\equiv0\pmod4$. These agree by Lemma~\ref{lem:invariant}, but the lemma as
usually stated gives only the forward implication, and it is the \emph{converse} the
search relies on: a word rejected for $\growth\not\equiv0$ must genuinely fail to
reset the heading, or a candidate would be silently skipped. For an arbitrary turn
word, the heading increment modulo $4$ is directly
$\#\turnR-\#\turnL=\growth$, so the two conditions are equivalent even before a
colour realisation is considered. We also confirmed the equivalence by enumeration over all
$524{,}286$ words of length at most $18$: the heading-resetting words number
$174{,}762$, and so do the words with $\growth\equiv0\pmod4$, with no exceptions in
either direction.

The residue theorem was audited separately, with the following counts over all
$\sum_{L=1}^{18}2^{L}=524{,}286$ nonempty turn words of length at most $18$:
$174{,}762$ restore the heading; $172{,}092$ of those have nonzero drift; and
\emph{none} of those $172{,}092$ passes the realisability criterion of
Theorem~\ref{thm:criterion}, consistent with the period-$\le48$ exclusion. The
residue theorem \eqref{eq:res-a}--\eqref{eq:P16} is a statement about \emph{realisable}
traces, so its direct test is vacuous in this length range; what the enumeration does
verify on all $172{,}092$ words is the underlying additive toggle-charge identity
\eqref{eq:conservation} (which holds for every heading-reset word, realisable or not).
The strand-level identity \eqref{eq:res-a} and the bound \eqref{eq:P16} are instead
confirmed on the standard word, its powers, quarter-turn rotations, and cyclic phase
shifts. (In
particular a word such as $\turnR\turnL$, with drift $(1,1)$ and growth $0$, restores
the heading and lies among the $172{,}092$ but is \emph{not} realisable and is
excluded from the strand-level test, as it must be.)

\subsection*{Environment and archived records}

The searches reported here were run on a single workstation, not on a cluster: one
Intel Core i5-1334U (10 physical cores, 12 logical, $1.3$\,GHz base) with $16$\,GB of
RAM under Windows~$11$ (build $26200$), on Eclipse Temurin OpenJDK $25.0.1{+}8$
(64-bit Server VM), with one JVM per shard and at most $12$ concurrent shards; the
Python verifier ran on CPython $3.13.14$. The period-$48$ run of the residue-free
variant is the largest single job: its shard records total $51{,}856.023$ worker-seconds
($14.40$ aggregate worker-hours). With at most $12$ one-JVM workers, the observed wall
time was on the order of one to two hours; the exact per-shard times are archived.

The engines' source, an SHA-256 manifest of the released snapshot, the aggregation
and cross-audit scripts, and the exact commands to regenerate every shard are publicly
available in the accompanying repository \cite{Repo}. For periods $34$--$48$, the
\emph{complete} per-rank shard records---each rank's node, leaf, prune, and
exact-evaluation counters, its wall time, and its hit list---are archived with the
paper rather than left to be regenerated. For period at most $32$, the archive instead
contains the unprefixed depth-$5$ baseline and all $32$ complete six-symbol prefix-shard
logs; these certify prefix coverage but do not contain per-rank subrecords. Compressed,
the full set is a few megabytes.
The archive also carries the aggregation output, the period-$34$--$48$ rank-coverage
verification (every rank covered exactly once), the
check that every configured cap was left unreached, the raw records needed for
per-rank comparison of the residue-free and pruned variants where both were run, the
exact commands, and the
environment strings above. The period-$34$--$48$ certifying shards have
\texttt{node\_cap=null}; the older period-$\le32$ shards record a $10^8$ safety cap
but all completed far below it. The exact search-record archive used for this revision
has SHA-256
\texttt{7052f3070829bf804ae317a40379f6bb045b00775aff826b4e347880747bcc1a}.

%=======================================================================
\begin{thebibliography}{99}

\bibitem{BunimovichTroubetzkoy1992}
L.~A. Bunimovich and S.~E. Troubetzkoy.
\newblock Recurrence properties of {L}orentz lattice gas cellular automata.
\newblock \emph{J. Stat. Phys.} \textbf{67} (1992), no.~1--2, 289--302.

\bibitem{Etse2026confined}
K.~R. Etse.
\newblock How long can the escaping ant be confined?
\newblock arXiv:2606.26677 (2026).

\bibitem{GajardoLutfallaRao2024}
A.~Gajardo, V.~H. Lutfalla, and M.~Rao.
\newblock Ants on the highway.
\newblock arXiv:2409.10124 (2024); \emph{Natural Computing} \textbf{24} (2025), 497--509.

\bibitem{GajardoMoreiraGoles2002}
A.~Gajardo, A.~Moreira, and E.~Goles.
\newblock Complexity of {L}angton's ant.
\newblock \emph{Discrete Appl. Math.} \textbf{117} (2002), no.~1--3, 41--50.
\newblock doi:10.1016/S0166-218X(00)00334-6; arXiv:nlin/0306022.

\bibitem{Gajardo2003symbolic}
A.~Gajardo.
\newblock A symbolic projection of {L}angton's ant.
\newblock \emph{Discrete Math. Theor. Comput. Sci. Proc.} AB (DMCS'03)
  (2003), 57--68.
\newblock doi:10.46298/dmtcs.2312.

\bibitem{GaleProppSutnerTroubetzkoy1995}
D.~Gale, J.~Propp, S.~Sutherland, and S.~Troubetzkoy.
\newblock Further travels with my ant.
\newblock \emph{Math. Intelligencer} \textbf{17} (1995), no.~3, 48--56.

\bibitem{Langton1986}
C.~G. Langton.
\newblock Studying artificial life with cellular automata.
\newblock \emph{Physica D} \textbf{22} (1986), no.~1--3, 120--149.
\newblock doi:10.1016/0167-2789(86)90237-X.

\bibitem{Lutfalla2025sideways}
V.~H. Lutfalla.
\newblock Sideways on the highways.
\newblock arXiv:2505.05426 (2025).

\bibitem{Repo}
A.~Jillhewar.
\newblock Langton's ant: exact searches, a Lean formalization, and supplementary data.
\newblock GitHub, \url{https://github.com/Atharva12456/langtons-ant-highway}; Zenodo
concept \textsc{doi}:\,\href{https://doi.org/10.5281/zenodo.21381637}{10.5281/zenodo.21381637}.
The search archive used here is pinned in Appendix~\ref{app:repro}; file-level hashes
are in \texttt{results/artifact\_hashes.json}.

\end{thebibliography}

\end{document}
